---
title: "popDE-Enrichment"
author: "Ruben Chazarra Gil"
date: "`r Sys.Date()`"
output: html_document
---

## 0. Goal 

Figure 3.  
- d) Pathway enrichment analysis for baseline NK cell population DTU genes (N=13; Hypergeometric test, FDR < 0.01). Gene Ratio indicates the fraction of DTU genes present in the enriched term gene set.

- Load functions
```{r}
source("../../../../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../../../../00_MetaData/03_Aesthetics-repo.R")
setwd("../03_Differential/02_DTU/02_DTU/02_popDTU/02_DA/")

suppressPackageStartupMessages(require(enrichplot)) # cannot install this in local!
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(scales))
# Run rrvgo
suppressPackageStartupMessages( require(rrvgo) )
suppressPackageStartupMessages(require(org.Hs.eg.db))
```


# Fig. 3. D

Figure 3.  
- d) Pathway enrichment analysis for baseline NK cell population DTU genes (N=13; Hypergeometric test, FDR < 0.01). Gene Ratio indicates the fraction of DTU genes present in the enriched term gene set.

```{r}
dtu.ora <- readRDS("04_popDTU-ORA/01_Processed-Data/06_celltype-fractions/01_popDTU-ORA-clusterProfiler-GO-BP-01_lineage.rds")

df <- dtu.ora$NK$NS$DTU.all$BP@result %>% filter(p.adjust < 0.01)
df <- df[order(df$p.adjust), ]
# n <- 20
# df <- df[1:n, ]
df$GeneRatio_numeric <- as.numeric(unlist(lapply(strsplit(df$GeneRatio, "/"), `[[`, 1))) / as.numeric(unlist(lapply(strsplit(df$GeneRatio, "/"), `[[`, 2)))
df$GeneRatio <- factor(df$GeneRatio, levels = sort(unique(df$GeneRatio)))
# Order by FDR inside each Gene Ratio
df <- df %>% arrange(GeneRatio, desc(p.adjust)) %>% mutate(Description = factor(Description, levels = unique(Description)))
df$Description <- factor(df$Description)

plt <- ggplot(data=df, aes(x = GeneRatio_numeric, y = Description, fill = p.adjust))+ 
  geom_point(shape=21, aes(size = Count), stroke = 0.1) + 
  scale_size_continuous("N Genes", range = c(1, 2))  + # shrink point sizes
  scale_fill_gradient(name = "FDR", low = "#023047", high = "#8ecae6") +
  scale_y_discrete(labels = function(x) stringr::str_trunc(x, 40) ) +
  scale_x_continuous(breaks =c(0.2, 0.3, 0.4, 0.5)) + 
  xlab("Gene Ratio") + # To be more informative "N of genes in enriched term in DTU set"
  plot_theme + theme(axis.line.x = element_line(linewidth=0.25), 
                     axis.title.x = element_text(size = 6), axis.title.y=element_blank()) + 
   guides(
    fill = guide_colorbar(order = 1, reverse =T),  # FDR legend comes first
    size = guide_legend(order = 2, override.aes = list(fill = "black"))     # Size legend comes second
  )
  
#  Save
file.name <- "04_popDTU-ORA/02_Plots/06_celltype-fractions/01_popDTU-ORA-01_lineage-NK_NS_DTU.all-DotPlot.pdf"
ggsave(file.name, plt, width = 6.25, height = 4.5, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, width = 6, height = 5, units = "cm", bg = "white")
```

