---
title: "Infection x population DTU interaction"
author: "Ruben Chazarra Gil"
date: "`r Sys.Date()`"
output: html_document
---

## 0. Goal 

*Figure 4. A*
a) Number of significant population × infection transcript usage (TU) interactions detected  (FDR<0.1), shown with and without adjustment for interindividual differences in cell composition.


- Load functions Functions
```{r}
source("../../../../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../../../../00_MetaData/03_Aesthetics-repo.R")
setwd("../03_Differential/02_DTU/02_DTU/03_interactionDTU/02_DA/")
```


# Figure 4.A

*Compare interaction DTU adjusting and no adjusting for cell type fractions*
```{r}
# 1. No adjusting
file.name <- "01_Processed-Data/05_Filt-1-5-0.8-0.01/01_interactionDTU-filtered-df-fdr-0.1.rds"
dtu.1 <- readRDS(file.name)$lineage
# 2. Adjusting
file.name <- "01_Processed-Data/06_celltype-fractions/01_interactionDTU-filtered-df-fdr-0.1.rds"
dtu.2 <- readRDS(file.name)$lineage

of.interest <- c("celltype", "Condition", "gene_name", "fdr")
dtu.1 <- dtu.1[ dtu.1$Condition %in% c("IAV.NS", "COV.NS"), of.interest ]
dtu.2 <- dtu.2[ dtu.2$Condition %in% c("IAV.NS", "COV.NS"), of.interest ]

dtu.1$Not_Adjusted <- T
dtu.2$Adjusted <-  T # "Adjusted"

df <- merge(dtu.1, dtu.2, by = c("celltype", "Condition", "gene_name"), all = T)
df[is.na(df)] <- F

table(df$Not_Adjusted, df$Adjusted)
df[df$Not_Adjusted & df$Adjusted, ]
```

- Plot numbers irrespective they are the same genes
```{r}
dtu1 <- readRDS("01_Processed-Data/05_Filt-1-5-0.8-0.01/01_interactionDTU-filtered-df-fdr-0.1.rds")
dtu2 <- readRDS("01_Processed-Data/06_celltype-fractions/01_interactionDTU-filtered-df-fdr-0.1.rds")

dtu1 <- dtu1$lineage %>% group_by(celltype, Condition) %>% summarise(n_genes = n_distinct(gene_id))
dtu2 <- dtu2$lineage %>% group_by(celltype, Condition) %>% summarise(n_genes = n_distinct(gene_id))

df <- merge(dtu1, dtu2, by = c("celltype", "Condition"), all = T)
names(df)[ match(c("n_genes.x", "n_genes.y"), names(df))] <- c("Not Adjusted", "Adjusted")
df <- df[df$Condition %in% c("IAV.NS", "COV.NS"), ]

df <- reshape2::melt(df, value.name = "n_genes", variable.name = "class")
df$class <- factor(df$class, levels = rev(c("Adjusted", "Not Adjusted")))

df$Condition <- factor(df$Condition, levels = names(condition.pairwise.vec.short))
df$Condition <- plyr::revalue(df$Condition, condition.pairwise.vec.short)
df$lineage <- df$celltype
df$lineage <- factor(df$lineage, levels = rev(lineage.df$lineage))

# Plot 
x_lab <- "N Infection * Population DTU genes"
plt <- ggplot(data = df, aes(y = lineage, x = n_genes, fill = lineage, alpha = class)) +
  geom_bar(stat = "identity", linewidth=0.25, position = position_dodge2(reverse = TRUE)) +
  facet_grid(Condition ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_fill_manual( values = lineage.color.vec ) +
  scale_alpha_manual("Cell composition", values = c(0.5, 1)) +
  xlab(x_lab) + ylab("y_lab") + 
  plot_theme + theme(axis.title.y = element_blank(), axis.title.x = element_text(hjust = 2), 
                     legend.position = "bottom") + 
  guides(fill = FALSE,  color = FALSE )
plt

# Save
file.name <- "02_Plots/06_celltype-fractions/01_01_interactDTU-N-DTU-01_lineage-Adjusted-vs-No-fdr-0.1.pdf"
ggsave(file.name, plt, width = 4.5, height = 5, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, width = 4.5, height = 5, units = "cm", bg = "white")
```

