---
title: "01_DA"
author: "Ruben Chazarra Gil"
date: "`r Sys.Date()`"
output: html_document
---

# 0. Goal

*DA and plot generation for TU-QTL section* 
```{r}
source("../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../../01_Randolph2021/05_Differential-Analyses/02_DTU/00_Scripts/00_MANTA-functions.R")
source("../../00_MetaData/03_Aesthetics-repo.R")
setwd("../06_sQTL/03_DA/")
```

# Response TU-QTL

*We define response TU QTLs as variants significantly associated with transcript usage only after infection (FDRIAV,COV < 0.01, pBaseline > 0.01)*

## Processing

```{r}
# Get Tested
df <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-tested-df.rds")
of.interest <- c("lineage", "Condition", "geneId", "gene_name", "snpId", "pv", "md")
df <- df[, of.interest]

# Add Significant FDR
sqtl <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
of.interest <- c("lineage", "Condition", "geneId", "gene_name", "snpId", "fdr")
sqtl <- sqtl[ , of.interest]
df <- merge(df, sqtl, by= c("lineage", "Condition", "geneId", "gene_name", "snpId"), all= T)

iav <- df[df$Condition %in% "IAV", ]
cov <- df[df$Condition %in% "COV", ]
ns <- df[df$Condition %in% "NS", ]
## Merge IAV & NS
iav.ns <- merge(iav, ns, by = c("lineage", "geneId", "gene_name", "snpId"), all.x = F, all.y = F)
revalue.vec <- c("Condition.x" = "Condition_1", "Condition.y" = "Condition_2", "md.x" = "MD.infected", "md.y" = "MD.baseline", "pv.x" = "p_value.infected", "pv.y" = "p_value.baseline", "fdr.x" = "fdr.infected", "fdr.y" = "fdr.baseline")
names(iav.ns) <- plyr::revalue(names(iav.ns), revalue.vec)
## Merge COV & NS
cov.ns <- merge(cov, ns, by = c("lineage", "geneId", "gene_name", "snpId"), all.x = F, all.y = F)
names(cov.ns) <- plyr::revalue(names(cov.ns), revalue.vec)

# Save tested response TU-QTLs: one row per SNP
df <- plyr::ldply(list("IAV_NS" = iav.ns, "COV_NS" = cov.ns), .id = "Condition")
saveRDS(df, "01_Processed-Data/03_celltype-fractions/02_00_response-TU-QTLs-tested.rds")

# Save significant response TU-QTLs
df <- df[!is.na(df$fdr.infected) & df$fdr.infected < 0.01 & is.na(df$fdr.baseline ) & df$p_value.baseline > 0.01, ] # Explanation: here we select those snps which are significant at the FDR level in infected but didn't make it to the FDR level in Baseline (response TU-QTLs)
saveRDS(df, "01_Processed-Data/03_celltype-fractions/02_01_response-TU-QTLs.rds")


# Save v2: save response tu QTLs present in either IAV_NS or COV_NS or both. Change formating for plots.
iav.cov <- merge(iav.ns, cov.ns, by = c("lineage", "geneId", "gene_name", "snpId"), all.x = F, all.y = F)
names(iav.cov) <- gsub("\\.x", ".IAV_NS", names(iav.cov))
names(iav.cov) <- gsub("\\.y", ".COV_NS", names(iav.cov))
# Add MD delta 
iav.cov$MD.delta.IAV_NS <- iav.cov$MD.infected.IAV_NS - iav.cov$MD.baseline.IAV_NS
iav.cov$MD.delta.COV_NS <- iav.cov$MD.infected.COV_NS - iav.cov$MD.baseline.COV_NS
# Keep response QTLs at least one of the Conditions
re.qtls.iav_ns <- iav.cov$fdr.infected.IAV_NS < 0.01 & iav.cov$p_value.baseline.IAV_NS > 0.01 
re.qtls.cov_ns <- iav.cov$fdr.infected.COV_NS < 0.01 & iav.cov$p_value.baseline.COV_NS > 0.01
# Deal with NAs
re.qtls.iav_ns[ is.na(re.qtls.iav_ns) ] <- F
re.qtls.cov_ns[ is.na(re.qtls.cov_ns) ] <- F
# Subset to re-tu-QTL in at least one Condition
iav.cov <- iav.cov[ re.qtls.iav_ns | re.qtls.cov_ns, ]
## add annotation 
iav.cov$re.TuQTL.IAV_NS <- ifelse(!is.na(iav.cov$fdr.infected.IAV_NS), T, F)
iav.cov$re.TuQTL.COV_NS <- ifelse(!is.na(iav.cov$fdr.infected.COV_NS), T, F)
saveRDS(iav.cov, "01_Processed-Data/03_celltype-fractions/02_02_response-TU-QTLs-IAV.NS-or-COV.NS.rds")
# save SNP ids for visualizations
write.table(iav.cov[, "snpId", drop = F], "01_Processed-Data/03_celltype-fractions/02_02_response-TU-QTLs-IAV.NS-or-COV.NS-SNP-ids.txt", quote = F, row.names=F, col.names =F)
```

<!-- ### Version 2  -->
<!-- ```{r} -->
<!-- # Get Tested -->
<!-- df <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-tested-df.rds") -->
<!-- of.interest <- c("lineage", "Condition", "geneId", "gene_name", "snpId", "pv", "md") -->
<!-- df <- df[, of.interest] -->

<!-- # Add Significant FDR -->
<!-- sqtl <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds") -->
<!-- of.interest <- c("lineage", "Condition", "geneId", "gene_name", "snpId", "fdr") -->
<!-- sqtl <- sqtl[ , of.interest] -->
<!-- df <- merge(df, sqtl, by= c("lineage", "Condition", "geneId", "gene_name", "snpId"), all= T) -->

<!-- iav <- df[df$Condition %in% "IAV", ] -->
<!-- cov <- df[df$Condition %in% "COV", ] -->
<!-- ns <- df[df$Condition %in% "NS", ] -->
<!-- ## Merge IAV & NS -->
<!-- iav.ns <- merge(iav, ns, by = c("lineage", "geneId", "gene_name", "snpId"), all.x = F, all.y = F) -->
<!-- revalue.vec <- c("Condition.x" = "Condition_1", "Condition.y" = "Condition_2", "md.x" = "MD.infected", "md.y" = "MD.baseline", "pv.x" = "p_value.infected", "pv.y" = "p_value.baseline", "fdr.x" = "fdr.infected", "fdr.y" = "fdr.baseline") -->
<!-- names(iav.ns) <- plyr::revalue(names(iav.ns), revalue.vec) -->
<!-- ## Merge COV & NS -->
<!-- cov.ns <- merge(cov, ns, by = c("lineage", "geneId", "gene_name", "snpId"), all.x = F, all.y = F) -->
<!-- names(cov.ns) <- plyr::revalue(names(cov.ns), revalue.vec) -->
<!-- # Save v1: one row per SNP -->
<!-- df <- plyr::ldply(list("IAV_NS" = iav.ns, "COV_NS" = cov.ns), .id = "Condition") -->
<!-- saveRDS(df, "01_Processed-Data/03_celltype-fractions/02_00_response-TU-QTLs-tested.rds") -->
<!-- df <- df[!is.na(df$fdr.infected) & df$fdr.infected < 0.01 & is.na(df$fdr.baseline ) & df$p_value.baseline > 0.01, ] # Explanation: here we select those snps which are significant at the FDR level in infected but didn't make it to the FDR level in Baseline (response TU-QTLs) -->
<!-- saveRDS(df, "01_Processed-Data/03_celltype-fractions/02_01_response-TU-QTLs.rds") -->
<!-- ``` -->
<!-- ``` -->

# N of re-TU-QTLs

```{r}
df <- readRDS( "01_Processed-Data/03_celltype-fractions/02_01_response-TU-QTLs.rds" )
df <- df %>% group_by(lineage, Condition) %>% summarise(n_genes = n_distinct(geneId)) 

# df$Condition <- gsub("_", ".", df$Condition)
# df$Condition <- factor(df$Condition, levels = rev(names(condition.pairwise.vec.short)))
# df$Condition <- plyr::revalue(df$Condition, condition.pairwise.vec.short)
df$Condition <- factor(df$Condition, levels = rev(c("IAV_NS", "COV_NS")))
df$Condition <- plyr::revalue(df$Condition, c("IAV_NS" = "IAV", "COV_NS" = "COV"))
df$lineage <- factor(df$lineage, levels = rev(lineage.df$lineage))

# Plot 
x_lab <- "Genes with a re-TU-QTL"
plt <- ggplot(data = df, aes(y = lineage, x = n_genes, fill = Condition)) +
  geom_bar(stat= "identity", linewidth=0.2, position = position_dodge()) + 
  scale_fill_manual( "Response", values = condition.color.vec.short ) +
  xlab(x_lab) + 
  plot_theme + theme(axis.title.y = element_blank(), 
                     legend.position=c(0.8, 0.45))+ 
  guides(fill = guide_legend(reverse = T))#, 
#+ guides(fill = FALSE,  color = FALSE,  alpha = guide_legend(override.aes = list(label = "", size = 1), nrow=2, reverse = T))

# 
file.name <- "02_Plots/03_celltype-fractions/04_01_N-response-TU-QTL-Genes.pdf"
ggsave(file.name, plt, width = 4, height = 4.25, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, width = 4, height = 4.25, units = "cm", bg = "white")
```


# Fig 4. C
*Condition sharing *
```{r}
df  <- readRDS("01_Processed-Data/03_celltype-fractions/02_00_response-TU-QTLs-tested.rds")
df$geneId_snpId <- paste0(df$geneId, "_", df$snpId)
# Split by lineage and Condition
df.l <- split(df, df$lineage)
df.l <- lapply(df.l, function(x) split(x, x$Condition))
# Subset to variants tested in all Conditions across lineages (To be fair we should have the separate intersection of IAV_Baseline and COV_Baseline ), in any case, we are being more restrictive
df.l <- lapply(df.l, function(c){
  vec = Reduce(intersect, lapply(c, function(x) x$geneId_snpId))
  print(length(vec))
  lapply(c, function(x) x[x$geneId_snpId %in% vec, ])
  })
df.l <- lapply(df.l, function(c) plyr::ldply(c, .id = "Condition"))
df <- plyr::ldply(df.l, .id = "lineage")

# Filter
df <- df[!is.na(df$fdr.infected) & df$fdr.infected < 0.01 & is.na(df$fdr.baseline ) & df$p_value.baseline > 0.01, ] # Explanation: here we select those snps which are significant at the FDR level in infected but didn't make it to the FDR level in Baseline (response TU-QTLs)

# Count sharing
df <- df %>% group_by(lineage, geneId_snpId) %>% summarise(n_conditions = n_distinct(Condition), Condition = paste(Condition, collapse = ","))
df <- df %>% group_by(lineage, Condition) %>% summarise(n_qtls = n_distinct(geneId_snpId))
df <- df %>% group_by(lineage) %>% mutate(n_qtls_rel = n_qtls /sum(n_qtls))
df
# Fraction of response TU-QTLs shared by IAV and COV
df[df$Condition %in% "IAV_NS,COV_NS", ]
# Fraction of response TU-QTLs spectifi to IAV or COV
df[df$Condition %in% c("IAV_NS", "COV_NS"), ]
# Fraction of response TU-QTLs spectifi to IAV and COV
df[df$Condition %in% c("IAV_NS", "COV_NS"), ] %>% group_by(lineage) %>% summarise(n_qtls_rel = sum(n_qtls_rel))
```

- Plot 
```{r}
df$Condition <- plyr::revalue(df$Condition, c("IAV_NS" = "IAV", "IAV_NS,COV_NS" = "Both", "COV_NS" = "COV"))
df$Condition <- factor(df$Condition, levels = rev(c("IAV", "Both", "COV")))
df$lineage <- factor(df$lineage, levels = rev(lineage.df$lineage))
color.vec <- c("Both" = "#f1faee", "IAV" = "#75A9D1", "COV"= "#c1121f")

# Plot 
x_lab <- "N of response TU-QTLs"
plt <- ggplot(data = df, aes(y = lineage, x = n_qtls, fill = Condition)) +
  #geom_bar(stat= "identity", linewidth=0.2, position = position_dodge()) + 
  geom_bar(stat= "identity", linewidth=0.2, position = "stack") + 
  scale_fill_manual( "Response", values = color.vec ) +
  xlab(x_lab) + 
  plot_theme + theme(axis.title.y = element_blank(), 
                     legend.position=c(0.8, 0.3))+ 
  guides(fill = guide_legend(reverse = T))#, 
#+ guides(fill = FALSE,  color = FALSE,  alpha = guide_legend(override.aes = list(label = "", size = 1), nrow=2, reverse = T))

# 
file.name <- "02_Plots/03_celltype-fractions/04_03_01_response-TU-QTL-Condition-sharing.pdf"
ggsave(file.name, plt, width = 4, height = 4.25, units = "cm")
# file.name <- gsub(".pdf", ".svg", file.name)
# ggsave(file.name, plt, width = 4, height = 4.25, units = "cm", bg = "white")
```


# Fig 4. E

*4.2 re-QTLs to IAV and COV Effect size scatter*
*Scatter plot of MD(IAV-Baseline) vs (COV-Baseline) for reQTLs*

- Correlate effect sizes
```{r}
df <- readRDS("01_Processed-Data/03_celltype-fractions/02_02_response-TU-QTLs-IAV.NS-or-COV.NS.rds")

# All
df %>% group_by(lineage) %>% summarise( cor = cor(MD.delta.IAV_NS, MD.delta.COV_NS, method = "pearson"), 
                                        p_value = cor.test(MD.delta.IAV_NS, MD.delta.COV_NS, method = "pearson")$p.value,.groups = "drop")

# Detected in both
length(unique(df[df$re.TuQTL.IAV_NS & df$re.TuQTL.COV_NS, ]))
cor.df <- df[df$re.TuQTL.IAV_NS & df$re.TuQTL.COV_NS, ] %>% group_by(lineage) %>% summarise( cor = cor(MD.delta.IAV_NS, MD.delta.COV_NS, method = "pearson"), 
                                                                                             p_value = cor.test(MD.delta.IAV_NS, MD.delta.COV_NS, method = "pearson")$p.value,.groups = "drop")
```

- Which are the STRONGEST response-TU-QTLs ? 
```{r}
df <- readRDS("01_Processed-Data/03_celltype-fractions/02_02_response-TU-QTLs-IAV.NS-or-COV.NS.rds")

df[which.max(df$MD.delta.IAV_NS), ]
df[which.max(df$MD.delta.COV_NS), ]
```

- Plot
```{r}
# Plot 
df <- readRDS( "01_Processed-Data/03_celltype-fractions/02_02_response-TU-QTLs-IAV.NS-or-COV.NS.rds" )
# Add whether re-TU-QTLs are in IAV or COV response or both
df$Condition <- paste0(df$re.TuQTL.IAV_NS, "_", df$re.TuQTL.COV_NS)
df$Condition <- plyr::revalue(df$Condition, c("FALSE_TRUE" = "COV", "TRUE_FALSE" = "IAV", "TRUE_TRUE" = "Both"))
color.vec <- c("Both" = "#e0e1dd", "IAV" = "#75A9D1", "COV"= "#c1121f")
df$Condition <- factor(df$Condition, levels = c("IAV", "COV", "Both"))

lineage.color.vec <- setNames(lineage.df$color, nm = lineage.df$lineage)
df$lineage <- factor(df$lineage, levels = names(color.vec))

plt <- ggplot(data = df, aes(x = MD.delta.COV_NS, y = MD.delta.IAV_NS, color = Condition)) + 
  xlim(-0.05, 0.12) + ylim(-0.05, 0.12) + 
  geom_point(size = 0.25) + 
  geom_vline(aes(xintercept = 0), linewidth = 0.25) +
  geom_hline(aes(yintercept = 0), linewidth = 0.25) +
  scale_color_manual("Response", values = color.vec ) +
  xlab("COV reTU-QTL effect size") + ylab("IAV reTU-QTL effect size") + 
  plot_theme + 
  guides(color = guide_legend(override.aes = list(size = 1)))

file.name <- paste0("02_Plots/03_celltype-fractions/04_02_response-TU-QTL_02-all-reQTLs-color-Condition.pdf")
ggsave(file.name, plt, width = 5.5, height = 4.5, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, width = 5.5, height = 4.5, units = "cm", bg = "white")
```


# Overlap reTU-QTLs with population * infection DTU interaction

*We found no significant overlap*

- popDRTU
```{r}
pop.l <-  readRDS("../../03_Differential/02_DTU/02_DTU/03_interactionDTU/02_DA/01_Processed-Data/06_celltype-fractions/01_interactionDTU-unfiltered-list.rds")
pop.l <- pop.l$lineage
pop.l <- lapply(pop.l, function(x) x[c("IAV.NS", "COV.NS")])

# v2 definition of interaction 
# pop <-  readRDS("../../03_Differential/02_DTU/02_DTU/03_interactionDTU/02_DA/01_Processed-Data/07_split-inf-baseline/01_interactionDTU-unfiltered-df.rds")
# pop.l <- split(pop, pop$celltype)
# pop.l <- lapply(pop.l, function(x) split(x, x$Condition))
```

- Our data
```{r}
df <- readRDS( "01_Processed-Data/03_celltype-fractions/02_00_response-TU-QTLs-tested.rds" )
# df <- readRDS( "01_Processed-Data/03_celltype-fractions/02_00_response-TU-QTLs-tested-v02.rds" )
df$Condition <- plyr::revalue(df$Condition, c("IAV_NS" = "IAV.NS", "COV_NS" = "COV.NS"))
df.l <- split(df, df$lineage)
df.l <- lapply(df.l, function(x) split(x, x$Condition))
```

- Overlap
```{r}
lineages <- names(pop.l)
conditions <- names(pop.l$B)

overlap.l <- lapply(lineages, function(ct) lapply(conditions, function(x) {
  print(paste0(ct, "_", x))
  df.1 <- df.l[[ct]][[x]]
  df.2 <- pop.l[[ct]][[x]]
  
  sig_1 <- unique(df.1[!is.na(df.1$fdr.infected) & df.1$fdr.infected < 0.01 & is.na(df.1$fdr.baseline ) & df.1$p_value.baseline > 0.01, "gene_name"])
  # sig_1 <- unique(df.1[ df.1$fdr.infected < 0.01 & df.1$fdr.infected < df.1$fdr.baseline & df.1$MD.infected > df.1$MD.baseline, "gene_name"])
  
  # Explanation: here we select those snps which are significant at the FDR level in infected but didn't make it to the FDR level in Baseline (response TU-QTLs)
  all_1 <- unique(df.1$gene_name)
  sig_2 <- df.2[df.2$fdr < 0.1, "gene_name", drop = T] # population * infection TU interaction at FDR < 0.1
  # sig_2 <- df.2[df.2$p.value < 0.1, "gene_name", drop = T] # population * infection TU interaction at p value < 0.1
  # sig_2 <- df.2[df.2$is.interaction, "gene_name" ]  # v2 definition of pop*infection TU interaction
  all_2 <- df.2[, "gene_name", drop = T] 
  
  df <- data.frame("shared" = length( intersect( sig_1, sig_2 )),
                   "model.1.specific" = length( setdiff( sig_1, sig_2 )), # model specific but SIG
                   "model.2.specific" = length( setdiff( sig_2, sig_1 )) )
  # contingency table
  cont.tab <- contingency.table.fun(sig_1, sig_2, all_1, all_2)
  ## add contingency table values
  df[["S12"]] <- cont.tab[1, 1] ; df[["S1_NS2"]] <- cont.tab[2, 1] ; df[["S2_NS1"]] <- cont.tab[1, 2] ; df[["NS12"]] <- cont.tab[2, 2]

  # Fisher exact test
  fisher_result <- fisher.test(cont.tab)
  df[["p_value"]] <- fisher_result$p.value
  df[["odds.ratio"]] <- fisher_result$estimate[["odds ratio"]]
  df[["conf.int.lower"]] <- fisher_result$conf.int[1]
  df[["conf.int.upper"]] <- fisher_result$conf.int[2]

  return(df)
}))

overlap.l <- lapply(overlap.l, function(ct) setNames(ct, conditions))
overlap.l <- setNames(overlap.l, lineages)

# Merge
overlap.l <- lapply(overlap.l, function(ct) plyr::ldply(ct, .id = "Condition"))
overlap.df <- plyr::ldply(overlap.l, .id = "lineage")
overlap.df$fdr <- p.adjust(overlap.df$p_value)
overlap.df
# saveRDS(overlap.df, "01_Processed-Data/03_celltype-fractions/02_04_05_response-TU-QTLs-overlap-popDRTU-split-metrics.rds")
```
# Supp. Fig. 21.

*Response TU-QTLs sharing Upset for: i) genes , ii ) QTLs*

- Read data 
```{r}
df <- readRDS( "01_Processed-Data/03_celltype-fractions/02_01_response-TU-QTLs.rds" )
df$Condition <- gsub("_", ".", df$Condition)
df$lineage_Condition <- paste0(df$lineage, "_", df$Condition)
df$geneId_snpId <- paste0(df$geneId, "_", df$snpId)
# Split by lineage_Condition unlist one level
df.l <- split(df, df$lineage_Condition)
# Get gene_id
# df.l <- lapply(df.l, function(x) x$geneId) # gene sharing
df.l <- lapply(df.l, function(x) x$geneId_snpId) # snpId sharing
# Pre-process for upset
df <- prepare.ComplexUpset(df.l)
```

- ComplexUpset Plot * Work has to be done here*
```{r}
# naming
# colnames(df) <-  gsub("(.*)\\.(.*\\..*)", "\\1_\\2", colnames(df))
name.df <- data.frame(ct_condition = colnames(df))
name.df$lineage <- gsub("_.*", "", name.df$ct_condition)
name.df$Condition <-  gsub(".*_", "", name.df$ct_condition)
name.df <- merge(name.df,  lineage.df[, c("lineage", "color")], by = "lineage")
name.df$Condition_2 <- plyr::revalue(name.df$Condition, condition.pairwise.vec.short)
name.df$lineage_Condition_2 <- paste0(name.df$lineage, " ", name.df$Condition_2)
# subset name.df
name.df <- name.df[name.df$ct_condition %in% colnames(df), ]
name.vec <- setNames(name.df$lineage_Condition_2, nm = name.df$ct_condition)

min_set_size = 1
font_size = 6
# Plot
plt <- ComplexUpset::upset(df, names(df), 
                           min_size = min_set_size,
                           set_sizes = FALSE, 
                           labeller = ggplot2::as_labeller(name.vec), 
                           name = NULL, 
                           height_ratio=0.8,
                           matrix=( intersection_matrix(geom=geom_point(size=1))), 
                           
                    base_annotations = list(
                      # Barplot
                      'N response TU-QTL' = intersection_size( fill = "#778da9",
                                                            text = list(size = 5/.pt, angle=90, vjust = 0.5, hjust = -0.2),
                                                            text_colors = c(on_background = "black",
                                                                            on_bar = "black")) +
                                      theme( panel.grid = element_blank(), axis.line = element_line(), 
                                             axis.title.y = element_text(size = font_size, vjust = -25),
                                             axis.text.y  = element_text(size = font_size))
                      ),
                    themes=upset_modify_themes(
                      list(
                        'intersection_size'=theme(text=element_text(size=font_size),
                                                  plot.margin = margin(0, 0, 0, 0, unit = "cm")),
                        'intersections_matrix'=theme(text=element_text(size=font_size),
                                                     plot.margin = margin(0, 0, 0, 0, unit = "cm"))
                        )), 
queries=list(upset_query(set= "T.CD4_IAV.NS", color = "#355070", fill="#355070"), 
             upset_query(set= "T.CD4_COV.NS", color = "#355070", fill="#355070"), 
             # upset_query(set= "T.CD4.IAV.COV", fill="#355070"), 
             upset_query(set= "T.CD8_IAV.NS", color = "#b56576", fill="#b56576"), 
             upset_query(set= "T.CD8_COV.NS", color = "#b56576", fill="#b56576"),
             # upset_query(set= "T.CD8_IAV.COV", color = "#b56576", fill="#b56576"),
             upset_query(set= "MONO_IAV.NS", color = "#6d597a", fill="#6d597a"), 
             upset_query(set= "MONO_COV.NS", color = "#6d597a", fill="#6d597a"),
             # upset_query(set= "MONO_IAV.COV", color = "#6d597a", fill="#6d597a"), 
             upset_query(set= "NK_IAV.NS", color = "#eaac8b", fill="#eaac8b"), 
             upset_query(set= "NK_COV.NS", color = "#eaac8b", fill="#eaac8b"),
             # upset_query(set= "NK.IAV.COV", fill="#eaac8b"), # no hits
             upset_query(set= "B_IAV.NS", color = "#e56b6f", fill="#e56b6f"),
             upset_query(set= "B_COV.NS", color = "#e56b6f", fill="#e56b6f")#,
             #upset_query(set= "B.IAV.COV", fill="#e56b6f") # no hits
             ))
plt

file.name<- paste0("02_Plots/03_celltype-fractions/05_01_responseTU-QTL-sharing-ComplexUpset-min-set-", min_set_size, "-QTLs.pdf")
ggsave(file.name, plt, width = 10, height = 5, units = "cm")
# file.name <- gsub(".pdf", ".svg", file.name)
# ggsave(file.name, plt, width = 12, height = 6, units = "cm", bg = "white")
```


# Supp Fig. 22 Overlap reTU-QTLs with population DTU

*Given we don't find reTU-QTL gene overlap with pop*infection interactions, let's try with standard popDTU*
- popDRTU
```{r}
pop.l <-  readRDS("../../03_Differential/02_DTU/02_DTU/02_popDTU/02_DA/01_Processed-Data/06_celltype-fractions/01_popDTU-unfiltered-list.rds")
pop.l <- pop.l$lineage
```

- Re TU-QTLs data
```{r}
df <- readRDS( "01_Processed-Data/03_celltype-fractions/02_00_response-TU-QTLs-tested.rds" )
# df$Condition <- plyr::revalue(df$Condition, c("IAV_NS" = "IAV", "COV_NS" = "COV"))
df.l <- split(df, df$lineage)
df.l <- lapply(df.l, function(x) split(x, x$Condition))
```

- Overlap
```{r}
lineages <- names(pop.l)
conditions.pop <- names(pop.l$B)
conditions.qlt <- names(df.l$B)

overlap.l <- lapply(lineages, function(ct) lapply(conditions.pop, function(x)
  lapply(conditions.qlt, function(y){
  df.1 <- df.l[[ct]][[y]]
  df.2 <- pop.l[[ct]][[x]]
  
  sig_1 <- unique(df.1[!is.na(df.1$fdr.infected) & df.1$fdr.infected < 0.01 & is.na(df.1$fdr.baseline ) & df.1$p_value.baseline > 0.01, "gene_name"])
  # sig_1 <- unique(df.1[ df.1$fdr.infected < 0.01 & df.1$fdr.infected < df.1$fdr.baseline & df.1$MD.infected > df.1$MD.baseline, "gene_name"])
  
  # Explanation: here we select those snps which are significant at the FDR level in infected but didn't make it to the FDR level in Baseline (response TU-QTLs)
  all_1 <- unique(df.1$gene_name)
  sig_2 <- df.2[df.2$fdr < 0.05, "gene_name", drop = T] # population * infection TU interaction at FDR < 0.1
  all_2 <- df.2[, "gene_name", drop = T] 
  
  df <- data.frame("shared" = length( intersect( sig_1, sig_2 )),
                   "model.1.specific" = length( setdiff( sig_1, sig_2 )), # model specific but SIG
                   "model.2.specific" = length( setdiff( sig_2, sig_1 )) )
  # contingency table
  cont.tab <- contingency.table.fun(sig_1, sig_2, all_1, all_2)
  ## add contingency table values
  df[["S12"]] <- cont.tab[1, 1] ; df[["S1_NS2"]] <- cont.tab[2, 1] ; df[["S2_NS1"]] <- cont.tab[1, 2] ; df[["NS12"]] <- cont.tab[2, 2]

  # Fisher exact test
  fisher_result <- fisher.test(cont.tab)
  df[["p_value"]] <- fisher_result$p.value
  df[["odds.ratio"]] <- fisher_result$estimate[["odds ratio"]]
  df[["conf.int.lower"]] <- fisher_result$conf.int[1]
  df[["conf.int.upper"]] <- fisher_result$conf.int[2]

  return(df)
  })))

overlap.l <- lapply(overlap.l, function(ct) lapply(ct, function(x) setNames(x, conditions.qlt)))
overlap.l <- lapply(overlap.l, function(ct) setNames(ct, conditions.pop))
overlap.l <- setNames(overlap.l, lineages)

# Merge
overlap.l <- lapply(overlap.l, function(ct) lapply(ct, function(x) plyr::ldply(x, .id = "Condition.QTL")))
overlap.l <- lapply(overlap.l, function(ct) plyr::ldply(ct, .id = "Condition.POP"))
overlap.df <- plyr::ldply(overlap.l, .id = "lineage")
overlap.df$fdr <- p.adjust(overlap.df$p_value)
overlap.df <- add.p_value.label(df = overlap.df, p_value.col="fdr")

overlap.df[overlap.df$fdr < 0.05, ]
overlap.df[overlap.df$Condition.POP == "NS", ]

# Plot
df = reshape2::melt(overlap.df[, c("lineage", "Condition.POP", "Condition.QTL",  "shared", "model.1.specific", "model.2.specific", "p_value_label")])
df$variable <- plyr::revalue(df$variable, c("model.1.specific" = "response TU-QTL", "model.2.specific" = "Population DTU"))
df$lineage <- factor(df$lineage, levels = rev(lineage.df$lineage))

df$Condition.POP <- factor(df$Condition.POP, levels = names(condition.vec.short))
df$Condition.POP <- plyr::revalue(df$Condition.POP, condition.vec.short)
df$Condition.QTL <- factor(df$Condition.QTL, levels = c("IAV_NS", "COV_NS"))
df$Condition.QTL <- plyr::revalue(df$Condition.QTL, c("IAV_NS" = "IAV-Baseline", "COV_NS" = "COV-Baseline"))

df <- df %>% group_by(lineage, Condition.POP, Condition.QTL) %>% mutate(value_rel = value/sum(value))
# Label df
label.df <- df %>% mutate(x = sum(value), x_rel = 1.02) %>% distinct(Condition.POP, Condition.QTL, p_value_label, x, x_rel) 
label.df[label.df$p_value_label == "ns", "p_value_label"] <- NA # replace "ns" with NA
label.df[!is.na(label.df$p_value_label), "p_value_label"] <- "*" # no need for multiplt "***" here
color.vec <- c("shared" = "#f2cc8f", "response TU-QTL" = "#81b29a", "Population DTU" = "#e07a5f")

# Plot 
plt <- ggplot(data = df, aes(y = lineage, x = value)) +
  geom_bar(aes( fill = variable), stat = "identity", linewidth=0.25, position = position_stack(reverse = T)) +
  geom_text(data = df[df$value > 0, ], aes(x = value, label = value), color = "black",  size = 5/.pt, hjust = 0.5, 
            position = position_stack(vjust = .5)) +
  # significance label
  geom_text(data = label.df, aes(x = x, y = lineage, label = p_value_label), color = "black",  size = 7/.pt, hjust = 0, vjust = 0.75) +
  facet_grid(Condition.POP ~ Condition.QTL, switch = "y") +
  scale_fill_manual("", values = color.vec ) +
  # scale_color_manual( values = lineage.color.vec ) +
  # scale_alpha_manual("", values = c(0.5, 1)) +
  xlab("N genes") + ylab("Population DTU") + 
  plot_theme + 
  guides( color = FALSE )
plt

file.name <- "02_Plots/03_celltype-fractions/04_06_response-TU-QTL-sharing-with-popDTU-01_absolute.pdf"
ggsave(file.name, plt, width =10.5, height = 7.5, units = "cm")
```


# Lineage sharing 

```{r}
df  <- readRDS("01_Processed-Data/03_celltype-fractions/02_00_response-TU-QTLs-tested.rds")
df$geneId_snpId <- paste0(df$geneId, "_", df$snpId)
# Split by Condition and lineage
df.l <- split(df, df$Condition)
df.l <- lapply(df.l, function(x) split(x, x$lineage))
# Subset to variatnts tested in all lineages
df.l <- lapply(df.l, function(c){
  vec = Reduce(intersect, lapply(c, function(x) x$geneId_snpId))
  print(length(vec))
  lapply(c, function(x) x[x$geneId_snpId %in% vec, ])
  })
df.l <- lapply(df.l, function(c) plyr::ldply(c, .id = "lineage"))
df <- plyr::ldply(df.l, .id = "Condition")
# Keep response QTLs
df <- df[!is.na(df$fdr.infected) & df$fdr.infected < 0.01 & is.na(df$fdr.baseline ) & df$p_value.baseline > 0.01, ] # Explanation: here we select those snps which are significant at the FDR level in infected but didn't make it to the FDR level in Baseline (response TU-QTLs)
# Count number of re-TU-QTLs and N of genes
# length(unique())

# Count sharing
df <- df %>% group_by(Condition, geneId_snpId) %>% summarise(n_lineages = n_distinct(lineage))
df <- df %>% group_by(Condition,  n_lineages) %>% summarise(n_qtls = n_distinct(geneId_snpId))
df <- df %>% group_by(Condition) %>% mutate(n_qtls_rel = n_qtls / sum(n_qtls) )
```

