---
title: "03_DTU-DA"
author: "Ruben Chazarra Gil"
date: "`r Sys.Date()`"
output: html_document
---

# 0. Goal 

- Infection DTU Data Analysis

Code to generate Figure 2. B, C, E, F, G

- Functions
```{r}
source("../../../../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../../../../00_MetaData/03_Aesthetics-repo.R")
setwd("../03_Differential/02_DTU/02_DTU/01_infectionDTU/02_DA") # come back to notebook 
```

# Figure 2. B

*Number of infection DTU genes upon influenza A virus (IAV) and SARS-CoV-2 (COV) stimulation across immune lineages.*^

```{r}
file.name <- "01_Processed-Data/05_Filt-1-5-0.8-0.01/01_infectionDTU-filtered-df.rds"
dtu.l <- readRDS(file.name)
dtu <- dtu.l$lineage
#  N of DTU genes
df <- dtu %>% group_by(Condition, celltype) %>% summarise(n_dtu = n_distinct(gene_id), .groups = "drop") 
# subset 
df <- df[df$Condition %in% c("IAV.NS", "COV.NS"), ]
df$Condition <- factor(df$Condition, levels = names(condition.pairwise.vec.short))
df$Condition <- plyr::revalue(df$Condition, condition.pairwise.vec.short)
df$celltype <- factor(df$celltype, levels = lineage.df$lineage)
color.vec <- setNames(lineage.df$color, nm = lineage.df$lineage)
model <- "rel_abundances ~ Intercept + Run + Age + POP"
# Plot 
# tit <- "infectionDTU Aquino 2023"
# sub_tit <- paste( model , " | FDR < 0.05")
# x_lab <- "Cell type"
y_lab <- "N infection DTU Genes"

# Plot 
plt <- ggplot(data = df, aes(x = celltype, y = n_dtu, fill = celltype )) +
  geom_bar(stat= "identity", linewidth=0.25, position = position_stack(), color = "white") + 
  # Labels
  geom_text(data = df, aes(label = n_dtu), color = "black", size = 3.5/.pt, hjust = 0.5, position = position_stack(vjust = .5)) +
  # xlim(0, 600) + 
  facet_grid( ~ Condition,  labeller = labeller(Condition = condition.pairwise.vec.short) ) +
  scale_fill_manual( values = lineage.color.vec ) +
  scale_color_manual("Condition", values = condition.pairwise.color.vec.short, drop = F) +
  # labs(title = tit, subtitle = sub_tit) + xlab(x_lab) +
  ylab(y_lab) + 
  plot_theme + 
  theme(axis.text.x = element_text(angle=30, hjust =1), axis.title.x = element_blank()) + 
  guides(fill = FALSE, color = FALSE ) 

plt
# Save
file.name <- "02_Plots/05_Filt-1-5-0.8-0.01/01_03_infectionDTU-01_lineage-by-Condition-Horizontal.pdf" ## Figure 2. B
ggsave(file.name, plt, width = 6, height = 3.5, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, width = 6, height = 3.5, units = "cm")
```


## Get numbers in the paper

"We observed that viral stimulation induces extensive remodeling of transcript usage across all immune lineages (1497 total unique infection DTU genes) (Fig. 2b; Supplementary Data 2a). The response was broader for IAV than for COV, with 47% more genes showing transcript usage shifts, consistent with the stronger transcriptional activation previously reported for IAV."

```{r}
# N unique infection DTU genes
file.name <- "01_Processed-Data/05_Filt-1-5-0.8-0.01/01_infectionDTU-filtered-df.rds"
dtu.l <- readRDS(file.name)
dtu <- dtu.l$lineage
# keep IAV-Baseline and COV-Baseline
dtu <- dtu[dtu$Condition %in% c("IAV.NS", "COV.NS"), ]
length(unique(dtu$gene_name))

# Comparison of responses
df <- dtu %>% group_by(Condition) %>% summarise(n_dtu_genes = length(fdr < 0.05))
# How much larger is IAV transcript usage response compared to COV response
(df[df$Condition %in% "IAV.NS", "n_dtu_genes"] / df[df$Condition %in% "COV.NS", "n_dtu_genes"] )-1
```

# Figure 2. C

Because immune activation is known to induce isoform switching25–28, we examined whether infection DTU genes showed changes in the dominant isoform. Following infection, 16–29% of DTU genes shift to an alternative dominant EC. Over half of these events (57.5%) correspond to switches in the most abundant isoform (Fig. 2c), as the switching ECs map to distinct transcript sets —for example, the ORMDL1 gene in B cells upon COV infection (Fig. 2d).

*How many of the EC switches involve ECs that of non overlapping transcripts? E.g t1 vs t1,t2 would overlap, t1 vs t2 would not*. This plot does not stratify by DTU response (e.g IAV-Baseline, COV-Baseline)
```{r}
file.name <- "01_Processed-Data/05_Filt-1-5-0.8-0.01/01_infectionDTU-filtered-df.rds"
dtu.l <- readRDS(file.name)
dtu <- dtu.l$lineage
dtu <- dtu[dtu$Condition %in% c("IAV.NS", "COV.NS"), ]
dtu$is.EC.switch <- as.logical(dtu$is.EC.switch)

#  N of Tx switches among EC switches 
df <- dtu %>% group_by(celltype, is.EC.switch, is.transcript.switch) %>% summarise(n_genes = n(), .groups = "drop")
df <- df %>% group_by(celltype) %>% mutate(n_genes_rel = n_genes / sum(n_genes))
df$celltype <- factor(df$celltype, levels = rev(lineage.df$lineage))
color.vec <- setNames(lineage.df$color, nm = lineage.df$lineage)
model <- "rel_abundances ~ Intercept + Run + Age + POP"
iso.switch.vec <- c("TRUE" = "Iso switch", "FALSE" = "No Iso switch" )
# Plot 
tit <- "infectionDTU Aquino 2023"
sub_tit <- paste( model , " | FDR < 0.05")
x_lab <- "% infection DTU genes with EC switch"
y_lab <- "Cell type"

# Plot 
df <- df[df$is.EC.switch, ]
plt <- ggplot(data = df, aes(x = n_genes_rel, y = celltype, fill = celltype, alpha = is.transcript.switch)) +
  geom_bar(aes(group = is.transcript.switch), stat= "identity", linewidth=0.25, position = position_stack(reverse = F), color = "white") + 
  scale_alpha_manual("", values = c(0.8, 1), labels = iso.switch.vec) +
  geom_text(data = df, aes(x = n_genes_rel, label = n_genes, group = is.transcript.switch), color = "black", alpha = 1,  # Labels
            size = 3.5/.pt, hjust = 0.5, position = position_stack(vjust = .5, reverse = F)) +
  scale_fill_manual( values = lineage.color.vec ) +
  labs(title = tit, subtitle = sub_tit) + xlab(x_lab) + ylab(y_lab) + 
  plot_theme + 
  theme(axis.title.y = element_blank(), axis.title.x = element_text(hjust = 1),
        legend.position="bottom", legend.box="horizontal", 
        legend.box.spacing = unit(0, "cm"),
        legend.box.margin = margin(0, 0, 0, 0, unit = "cm"),
        plot.margin = unit(c(0.1, 0.1, 0, 0.1), "cm"),
        legend.spacing.y = unit(0, "cm"), 
        legend.key.spacing.x = unit(0, "cm")) + 
  guides(fill = FALSE, color = FALSE, 
          alpha=guide_legend( override.aes = list(label = "", size = 1, ncol=1), reverse = T)) 

plt
# Save
file.name <- "02_Plots/05_Filt-1-5-0.8-0.01/01_07_02_infectionDTU-N-DTU-EC-switch-Transcript-switch-no-Condition.pdf"
ggsave(file.name, plt, width = 4, height = 3.5, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, width = 4, height = 3.5, units = "cm", bg = "white")
```

## Get numbers in the paper
- Numbers in paper
```{r}
# How many infection DTU genes have a switch in their most abundant EC ? ("EC switch")
file.name <- "01_Processed-Data/05_Filt-1-5-0.8-0.01/01_infectionDTU-filtered-df.rds"
dtu.l <- readRDS(file.name)
dtu <- dtu.l$lineage
dtu <- dtu[dtu$Condition %in% c("COV.NS", "IAV.NS"), ]
df <- dtu %>% group_by(celltype, Condition, is.EC.switch) %>% summarise(n_genes = n())
df <- df %>% group_by(celltype, Condition) %>%  mutate (n_genes_rel = n_genes / sum(n_genes))
df <- df[df$is.EC.switch == "TRUE", ]
df[ which.min(df$n_genes_rel), ]
df[ which.max(df$n_genes_rel), ]

# Within EC switches, how many are "isoform switches", given the switching ECs map to distinct transcript sets (e.g, EC1{t1,t2} is dominant in Baseline, and EC2{t3} is dominant in IAV)
file.name <- "01_Processed-Data/05_Filt-1-5-0.8-0.01/01_infectionDTU-filtered-df.rds"
dtu.l <- readRDS(file.name)
dtu <- dtu.l$lineage
dtu <- dtu[dtu$Condition %in% c("COV.NS", "IAV.NS"), ]
dtu <- dtu[dtu$is.EC.switch == "TRUE", ]
df <- dtu %>% group_by(is.transcript.switch) %>% summarise(n_genes = n())
df <- df %>%  mutate (n_genes_rel = n_genes / sum(n_genes))
```

# Fig 2.E

*Fig 2.E* Heatmap of infection DTU gene sharing based on the spearman correlation of MD values considering genes significant in at least one of the lineage-response combinations. Hierarchical clustering of lineage-response pairs is displayed.

```{r}
library(RColorBrewer)

dtu.l <- readRDS("01_Processed-Data/05_Filt-1-5-0.8-0.01/01_infectionDTU-unfiltered-list.rds")
dtu.l <- dtu.l$lineage
dtu.l <- lapply(dtu.l, function(x) x[c("IAV.NS", "COV.NS")])
dtu.l <- unlist(dtu.l, recursive = F)
gene.l <- lapply(dtu.l, function(x) x$gene_name)
# Get intersection of genes tested in all conditions
# gene.vec <- Reduce(intersect, gene.l) 
# dtu.l <- lapply(dtu.l, function(x){  x[match(gene.vec, x$gene_name),]}) # subset

cor.l <- lapply(dtu.l, function(x1) # correlate MDs
  lapply(dtu.l, function(x2){
    print(paste0(nrow(x1), ":", nrow(x2)))    
     df <- merge(x1, x2, by = c("gene_id", "gene_name"), all = F)
     # keep infection DTU genes significant in at least one condition 
     keep.genes <- df$is.DTU.lineage.x | df$is.DTU.lineage.y
     df <- df[ keep.genes , ]
     cor.result <- cor.test(df$fdr.x, df$fdr.y, method = "spearman")
     data.frame(cor = cor.result[["estimate"]][["rho"]], p.val = cor.result$p.value)
  }))

# Create matrix
l <- lapply(cor.l, function(c_1) plyr::ldply(c_1, .id = "celltype_1"))
df <-  plyr::ldply(l, .id = "celltype_2")
cts <- unique(df$celltype_1)
n_cts <- length(cts)
mat <- matrix(df$cor, nrow = n_cts, ncol = n_cts, dimnames = list(cts, cts))

# Annotation and annotation colors 
annotation_col <- data.frame(lineage = gsub(".IAV.NS|.COV.NS", "", rownames(mat)),
                             Condition = gsub("B.|NK.|MONO.|T.CD4.|T.CD8.", "", rownames(mat))
)
annotation_col$Condition <- gsub(".NS", "", annotation_col$Condition) # shorten condition
# Change mat names
new_names <- paste0(annotation_col$lineage, " ", annotation_col$Condition)
rownames(mat) <- colnames(mat) <- new_names
rownames(annotation_col) <- rownames(mat)  # match column names of mat

ann_colors <- list(lineage = lineage.color.vec, Condition = c( "IAV" = "#75A9D1", "COV" = "#c1121f") )

# Define Spectral palette (reversed so blue → red)
spectral_palette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))(16)

# Heatmap with lineage and Condition annotations
ph <- pheatmap::pheatmap(
  mat = mat,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  color = spectral_palette,
  display_numbers = FALSE,
  annotation_col = annotation_col,
  annotation_row = annotation_col,
  annotation_colors = ann_colors,
  legend = TRUE,
  legend_labels = as.character(seq(-0.4, 1, 0.2)),  # Add legend title
  legend_breaks = seq(-0.4, 1, 0.2),
  main = NA,  # Remove plot title
  fontsize = 9, # to adjust for inch saving ? 
  border_color = "white"
)

# Save
file.name <- paste0("02_Plots/05_Filt-1-5-0.8-0.01/01_06_04-MD-clustering-01_lineage.pdf")  ## Fig 2.E
inch = 2.54
pdf(file.name, height = 11/inch, width = 14/inch)
print(ph)
dev.off()
```


### Figure 2. F

*Condition sharing of infection DTU genes*

f) Sharing of infection DTU genes across IAV and COV responses considering genes tested in both responses for each immune lineage separately.

```{r}
file.name <- "01_Processed-Data/05_Filt-1-5-0.8-0.01/01_infectionDTU-unfiltered-list.rds"
dtu.l <- readRDS(file.name)
dtu <- dtu.l$lineage
dtu <- list("IAV.NS" = lapply(dtu, function(ct) ct[["IAV.NS"]]), "COV.NS" = lapply(dtu, function(ct) ct[["COV.NS"]]))

dtu <-lapply(dtu, function(ct){ # subset for genes tested in IAV and COV across immune lineages
  gene.l <- lapply(ct, function(x) x$gene_id)
  gene.vec <- Reduce(intersect, gene.l)
  ct <- lapply(ct, function(x) x[x$gene_id %in% gene.vec , ] )
  data.table::rbindlist(ct) # merge
})
df <- data.table::rbindlist(dtu)
df <- df %>% group_by(Condition, gene_name) %>% filter(fdr < 0.05) %>% summarise(n_lineages = n_distinct(lineage))
df <- df %>% group_by(n_lineages) %>% summarise(n_genes = n()) # irrespective of condition
df <- df %>%  mutate(n_genes_rel = n_genes / sum(n_genes)) 
df$n_lineages <- factor(df$n_lineages, levels = c(1:5))

plt <- ggplot(df, aes(y = NA, x = n_genes_rel, fill = n_lineages)) + 
  geom_bar(stat = "identity", color = "black", linewidth = 0.25, position = position_stack(reverse = T)) + 
   geom_text(data = df[df$n_lineages %in% c("1", "2"), ], 
             aes(label = round(n_genes_rel, 2)), position = position_stack(vjust = 0.5, reverse = T), color="black", size=3.5/.pt) + 
  scale_fill_brewer("N lineages", palette = "Blues") + 
  xlab("% of infection DTU Genes") + 
  plot_theme + 
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(), 
        axis.line.y = element_blank())+ 
  guides(fill = guide_legend(byrow = FALSE, override.aes = list(color = "white", linewidth = 0.15)))

plt

lin.plt <- plt
# file.name <- "02_Plots/05_Filt-1-5-0.8-0.01/04_02_infectionDTU-sharing-BarPlot-no-Condition.pdf"
# ggsave(file.name, plt, width = 5, height = 2.5, units = "cm")
```

# Figure 2. G

*Lineage  sharing of infection DTU genes*

g) Sharing of infection DTU genes across immune lineages considering genes tested in all lineages for each response separately.

```{r}
file.name <- "01_Processed-Data/05_Filt-1-5-0.8-0.01/01_infectionDTU-unfiltered-list.rds"
dtu.l <- readRDS(file.name)
dtu <- dtu.l$lineage
dtu <-lapply(dtu, function(ct){ # subset for genes tested in IAV and COV across cell types
  ct <- ct[c("IAV.NS", "COV.NS")]
  gene.l <- lapply(ct, function(x) x$gene_id)
  gene.vec <- Reduce(intersect, gene.l)
  ct <- lapply(ct, function(x) x[x$gene_id %in% gene.vec, ] )
  data.table::rbindlist(ct) # merge
})
df <- data.table::rbindlist(dtu) # merge by lineage

df <- df %>% group_by(lineage, gene_name) %>% filter(fdr < 0.05) %>% summarise(Condition = paste(Condition, collapse = ","))
df <- df %>% group_by(Condition) %>% summarise(n_genes = n()) # irrespective of lineage, counting all lineages
df <- df %>% mutate(n_genes_rel = n_genes / sum(n_genes) )

df$Condition <- plyr::revalue(df$Condition, c("COV.NS" = "COV", "COV.NS,IAV.NS" = "Both", "IAV.NS,COV.NS" = "Both", "IAV.NS" = "IAV"))
df$Condition <- factor(df$Condition, levels = c("COV", "Both", "IAV"))
color.vec <- c("IAV" = "#75A9D1", "Both" = "#f1faee",  "COV"= "#c1121f")
df$Condition <- factor(df$Condition, levels = names(color.vec))

plt <- ggplot(df, aes(y = NA, x = n_genes_rel, fill = Condition)) + 
  geom_bar(width=0.8, stat = "identity", color = "black", 
           linewidth = 0.25, position = position_stack(reverse = T)) + 
  geom_text(aes(label = round(n_genes_rel, 2)), position = position_stack(vjust = 0.5, reverse = T), color="black", alpha = 1, size=3.5/.pt, angle = 0) + 
  scale_alpha_discrete("", range=c(1, 0.5, 0.25)) +
  scale_fill_manual( values = color.vec ) +
  xlab("% of infection DTU Genes") + 
  plot_theme + 
  theme(axis.title.y = element_blank(), axis.line.y = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        legend.box.spacing = unit(0, "cm"),
        legend.key.size = unit(0.1, "cm"),
        legend.title = element_text(size = 5, margin = margin(b = 0, unit = "cm")),
        legend.justification = c(1, 1), 
        legend.key = element_rect(fill = "white", color = NA) 
        ) + 
   guides(fill = guide_legend(byrow = FALSE, override.aes = list(color = "white", linewidth = 0.15)))

plt
cond.plt <- plt
file.name <- "02_Plots/05_Filt-1-5-0.8-0.01/04_03_infectionDTU-Condition-sharing.pdf"
ggsave(file.name, plt, width = 5, height = 1.5, units = "cm")
```

- Merge Condition No DGE + Lineage No DGE
```{r}
a <- cowplot::plot_grid(
    # 1) N genes
    cond.plt + theme (axis.text.x = element_blank(), 
                      axis.title.x = element_blank(), 
                      axis.line.x = element_blank(), 
                      axis.ticks.x = element_blank(), 
                      legend.justification = c(0, 1)), 
    # 2) Sharing barplot
    lin.plt,# + theme(legend.justification = c(-2, 1.5)),  
    ncol = 1, align = "v", axis = "rl" , rel_heights =  c(0.3, 0.7)
    )

file.name <- "02_Plots/05_Filt-1-5-0.8-0.01/04_03_02_infectionDTU-Lineage-Condition-sharing.pdf"
ggsave(file.name, a, width = 5, height = 3.5, units = "cm")
```