---
title: "01_DA"
author: "Ruben Chazarra Gil"
date: "`r Sys.Date()`"
output: html_document
---

# 0. Goal

Figure 3.  
- f) Number of genes with a TU-QTL across lineages and conditions. 
- g) Enrichment in TU-QTLs among population DTU genes. For each immune lineage and condition, the bars indicate the percentage of genes with a significant TU-QTL among population DTU tested and significant genes, respectively. 
- h) Heatmap of TU-QTLs sharing based on the spearman correlation of MD values considering gene-SNP pairs significant in at least one of the lineage-condition comparisons. In addition, hierarchical clustering of lineage-condition pairs is displayed.



- Load functions
```{r}
source("../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../../01_Randolph2021/05_Differential-Analyses/02_DTU/00_Scripts/00_MANTA-functions.R")
source("../../00_MetaData/03_Aesthetics-repo.R")
setwd("../06_sQTL/03_DA/")
```


# ADD to 0.2 Pre-processing
*In 00_collect-sQTLseekeR2-results.R* 

# Fig. 3.F

*Figure 3.f) Number of genes with a TU-QTL across lineages and conditions. *

- Get numbers in paper
```{r}
df = readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
count.df <- df %>% group_by(lineage, Condition) %>% summarise(n_genes = n_distinct(geneId), n_snps = n_distinct(snpId))
# N SNPs
count.df[which.min(count.df$n_snps), ]
count.df[which.max(count.df$n_snps), ]
# N genes
count.df[which.min(count.df$n_genes), ]
count.df[which.max(count.df$n_genes), ]
```


- Plot
```{r}
df <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
df <- df %>% group_by(lineage, Condition) %>% summarise(n_genes = n_distinct(geneId))

df$Condition <- factor(df$Condition, levels = names(condition.vec.short) )
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)
color.vec <- setNames(lineage.df$color, nm = lineage.df$lineage)
df$lineage <- factor(df$lineage, levels = rev(names(color.vec)))

# Plot 
x_lab <- "N Genes with a TU-QTL"

# Plot 
plt <- ggplot(data = df, aes(x = n_genes, y = lineage, fill = lineage, color = lineage)) +
  geom_bar(stat= "identity", linewidth=0.25) + 
  geom_text(data = df, aes(x = n_genes, label = n_genes), color = "black",  size = 5/.pt, hjust = 0.5, position = position_stack(vjust = .5)) +
  facet_grid(Condition ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_fill_manual( values = lineage.color.vec ) +
  scale_color_manual( values = lineage.color.vec ) +
   xlab(x_lab) + 
  plot_theme + theme(axis.title.y = element_blank()) + 
  guides(fill = FALSE,  color = FALSE )
plt

# # Merge with N cells per lineage 
# plt <- cowplot::plot_grid(
#     # 1) N cells / cell type
#     n_cell.lineage.by.condition.plt,
#     # 2) N tested genes
#     plt +  theme(strip.text.y = element_blank(), strip.background = element_blank(),
#                  axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank()),
#     nrow = 1, align = "h", axis = "tb" , rel_widths =  c(0.5, 0.5)
#     )

# Save
file.name <- "02_Plots/03_celltype-fractions/01_01_N-TU-Genes-by-Condtion.pdf"
ggsave(file.name, plt, width = 4.5, height = 5, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, width = 4.5, height = 5, units = "cm", bg = "white")
```

# Fig. 3.G

*Figure 3.  g) Enrichment in TU-QTLs among population DTU genes. For each immune lineage and condition, the bars indicate the percentage of genes with a significant TU-QTL among population DTU tested and significant genes, respectively.*

## Plot

*% of popDTU genes with a TU QTL and % of all tested with a TU QTL*
```{r}
# Load QTLs
qtl.sig <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
qtl.sig <- qtl.sig %>% distinct(lineage, Condition, gene_name)
qtl.sig$is.QTL <- T

# PopDTU tested
dtu.all <- readRDS("../../03_Differential/02_DTU/02_DTU/02_popDTU/02_DA/01_Processed-Data/06_celltype-fractions/01_popDTU-unfiltered-df.rds")
dtu.all <- dtu.all$lineage
dtu.all$lineage <- dtu.all$celltype
dtu.all <- dtu.all %>% distinct(lineage, Condition, gene_name, fdr)
dtu.all$is.DTU <- ifelse(dtu.all$fdr < 0.05,  T, F)
# Merge 
df <- merge(dtu.all, qtl.sig, by = c("lineage", "Condition", "gene_name"), all.x = T, all.y = F)
df[is.na(df$is.QTL), "is.QTL"] <- F

# Count how many of popDTU and non popDTU have a TU-QTL
df <- df %>% group_by(lineage, Condition, is.DTU, is.QTL) %>% summarise(n_genes = n_distinct(gene_name))
df <- df %>% group_by(lineage, Condition, is.DTU) %>% mutate( n_genes_rel = n_genes / sum(n_genes)) 
# We are interested in those genes with a QTL
df <- df[df$is.QTL, ]
# new label 
df$class <- ifelse(df$is.DTU & df$is.QTL, "popDTU", "Tested")

df$Condition <- factor(df$Condition, levels = names(condition.vec.short) )
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)
color.vec <- setNames(lineage.df$color, nm = rev(lineage.df$lineage))
df$lineage <- factor(df$lineage, levels = names(color.vec))

# Plot 
y_lab <- "N of TU-Genes"

# Plot 
plt <- ggplot(data = df, aes(y = lineage, x = n_genes_rel, fill = lineage, alpha = class)) +
  geom_bar(stat= "identity", linewidth=0.2, position = position_dodge()) + 
  scale_alpha_manual("", values = c(0.5, 1)) +
  # geom_text(data = df, aes(y = shared.rel.1, label = shared), # Text labels
  #           color = "black", alpha = 1,  size = 5/.pt, hjust = 0.5, position = position_stack(vjust = .5)) +
  # geom_text(aes(x = lineage, y = label_y, label = p_value_label), vjust = 0, size = 1.75) + 
  facet_grid( Condition ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_fill_manual( values = lineage.color.vec ) +
  xlab("Genes with a TU-QTL (%)") + 
  plot_theme + theme(axis.title.y = element_blank()#, , 
                     #axis.text.x = element_text(angle = 30, vjust = 0.5, hjust = 0.5), 
                     #legend.position = "bottom"
                     ) + 
  guides(fill = FALSE,  color = FALSE, 
         alpha = guide_legend(override.aes = list(label = "", size = 1, ncol=1), reverse = T))

# 
file.name <- paste0("02_Plots/03_celltype-fractions/03_01_Overlaps-PopDTU-02_relative-count-02_Genome-wide.pdf")
ggsave(file.name, plt, width = 5, height = 5, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, width = 5, height = 5, units = "cm", bg = "white")
```

## Overlap statistics of TU-QTL genes with popDTU genes

- DTU genes
```{r}
dtu.l <- readRDS("../../03_Differential/02_DTU/02_DTU/02_popDTU/02_DA/01_Processed-Data/06_celltype-fractions/01_popDTU-unfiltered-list.rds")
dtu.l <- dtu.l$lineage
```

- Process QTLs
```{r}
qtl.sig <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
qtl.sig <- qtl.sig %>% distinct(lineage, Condition, gene_name, fdr)
# add tested genes
qtls.all <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-tested-df.rds")
# colnames(qtls.all) <- plyr::revalue(colnames(qtls.all), c("cell.lineage" = "lineage", "condition" = "Condition"))
# qtls.all$Condition <- plyr::revalue(qtls.all$Condition, c("Baseline" = "NS", "Influenza A Virus" = "IAV", "SARS-CoV-2" = "COV"))
qtls.all <- qtls.all %>% distinct(lineage, Condition, gene_name)
qtl <- merge(qtl.sig, qtls.all, by = c("lineage", "Condition", "gene_name"), all = T)
qtl[ is.na(qtl$fdr), "fdr"] <- 1 # add 1 as FDR for NS gene-variant pairs
# Split
qtl.l <- split(qtl, qtl$lineage)
qtl.l <- lapply(qtl.l, function(x) split(x, x$Condition))
```

- Overlap
```{r}
# check
names(dtu.l) == names(qtl.l)

ov.l <- test.overlap.pro( m1 = dtu.l,
                  m2 = qtl.l,
                  fdr.col.1 = "fdr", fdr.col.2 = "fdr",
                  fdr.thres.1 = 0.05, fdr.thres.2 = 0.01,
                  gene.col.1 = "gene_name", gene.col.2 = "gene_name", 
                  FC.col.1 = NA, FC.col.2 = NA) 
# merge by Condition 
ov.l <- lapply(ov.l, function(x) plyr::ldply(x, .id = "Condition"))
ov.df <- plyr::ldply(ov.l, .id = "lineage")
ov.df <- add.p_value.label(ov.df, p_value.col = "p_value")
# add relative counts
df <- ov.df
df.l <- lapply(c(1:nrow(df)), function(i){
  x = df[i, ]
  
  model.1.total <- x[["model.1.specific"]] + x[["shared"]]
  model.2.total <- x[["model.2.specific"]] + x[["shared"]]
  
  x[["model.1.specific.rel"]] <- x[["model.1.specific"]] / model.1.total
  x[["model.2.specific.rel"]] <- x[["model.2.specific"]] / model.2.total
  x[["shared.rel.1"]] <- x[["shared"]] / model.1.total
  x[["shared.rel.2"]] <- x[["shared"]] / model.2.total
  return(x)
})

df.i <- data.table::rbindlist(df.l)
# add FDR
df.i$fdr <- p.adjust(df.i$p_value)
```


# Fig 3.H 

*Figure 3. h) Heatmap of TU-QTLs sharing based on the spearman correlation of MD values considering gene-SNP pairs significant in at least one of the lineage-condition comparisons. In addition, hierarchical clustering of lineage-condition pairs is displayed.*
