---
title: "04_popDTU-Validation"
author: "Ruben Chazarra Gil"
date: "`r Sys.Date()`"
output: html_document
---


# 0. Goal 

*Validation of our infection DTU genes with external datasets*

Supplementary Fig. 4. Validation of infection-associated differential transcript usage (DTU) detection by permutation analysis and overlap with previous studies. a) Permutation analysis for infection-associated DTU detection. Under the original (non-permuted) infection DTU analysis, in each viral response (e.g. influenza A virus (IAV)-Baseline, SARS-CoV-2 (COV)-Baseline), equivalence class (EC) differences were computed between matched infected and baseline samples from the same donor. (e.g. Donor1 Infected-Donor1 Baseline,  Donor2 Infected-Donor2 Baseline). Here, instead, we randomly shuffled the condition labels across donors, such that EC ratio differences were calculated between mismatched infected and baseline samples (e.g., Donor1 Infected-Donor1 Baseline,  Donor2 Baseline-Donor2 Infected). This procedure was repeated 100 times, and the infection DTU analysis was performed for each permutation. Boxplots show the number of significant DTU genes detected per permutation (center line: median; box limits: upper and lower quartiles; whiskers: 1.5× interquartile range; points: outliers). *b) Overlap with previously reported infection-associated splicing differences. Overlap between monocyte DTU genes identified in our study following IAV and COV infection and those reported in a bulk RNA-seq study of IAV-infected monocytes⁴. Enrichment is shown as odds ratio, with the dotted line indicating no enrichment (odds ratio = 1). Asterisks denote significance levels (* p < 0.01, ** p < 0.001, etc.).*


- Load Functions
```{r}
source("../../../../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../../../../00_MetaData/03_Aesthetics-repo.R")
setwd("../03_Differential/02_DTU/02_DTU/01_infectionDTU/02_DA") # come back to notebook 
```

# 2. Overlaps with external datasets

## 2.1 Rotival 2019 (Nature Communications, 2019)

`Rotival, M., Quach, H. & Quintana-Murci, L. Defining the genetic and evolutionary architecture of alternative splicing in response to infection. Nat Commun 10, 1671 (2019). https://doi.org/10.1038/s41467-019-09689-7`

```{r}
source("../../../../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../../../../../01_Randolph2021/05_Differential-Analyses/02_DTU/00_Scripts/00_MANTA-functions.R")
```

### 2.1.1 Processing

NOTE: 
(*Rotival 2019 > Methods > Differential splicing and isoform diversity*) Rotival considers differential splicing between stimulation conditions or populations: 
1) Only genes with FPKM > 10, both at basal state and after stimulation were considered (not considered labelled with "n.d" in Supp Data 4)
2) FDR < 0.05
3) Difference in PSI > 0.05
4) We considered genes as being differentially spliced (between conditions or populations) when they signiﬁcantly changed their isoform ratios (PSI), for at least one of their AS events
Misc: 
5) 2025-06-02: Rotival2019 models FC of Baseline / condition, hence "-" means upregulation in condition, and "+" means upregulation in baseline.

- Load Supplementary Data 2
```{r}
suppressPackageStartupMessages(require(readxl))
file.name <- "../../../../../../00_Bibliography/2019-Rotival-Nat_splicing-response-to-infection-Supp/2019-Rotival-Nat_splicing-response-to-infection-SupplementaryData_2.xlsx"
rotival <- read_excel(file.name, sheet = 2, skip = 13)
str(rotival)

#  Select data of interest: *NI* & *IAV infected*
## manually determined by looking at excel sheet
common.cols <- names(rotival)[ c(1,2, 3, 5, 6 ) ]

# Comparisons are all Not-Stimullated vs Stimullations

# # 1) Non-Stimullated

# 2) LPS-Stimullated
lps.cols <- names(rotival)[c(7:11, 28)]
lps <- rotival[ , c(common.cols, lps.cols) ]
# 3) PAM
pam.cols <- names(rotival)[c(12:16, 28)]
pam <- rotival[ , c(common.cols, pam.cols) ]
# 4) R8
r8.cols <- names(rotival)[c(17:21, 29)]
r8 <- rotival[ , c(common.cols, r8.cols) ]
# 5) IAV-Stimullated
iav.cols <- names(rotival)[c(22:26, 30)]
iav <- rotival[, c(common.cols, iav.cols) ]
## collect to list
rotival.l <- list("NS.LPS" = lps, "NS.PAM" = pam, "NS.R8" = r8, "NS.IAV" = iav)

# update names
update.names <- c("AS_event_id", "gene_symbol", "event_type", "Is_DS_any_condition", "AS_NS",
          "AS", "Delta_psi_NS_Stimullation", "p_value_NS_Stimullation", "FDR_NS_Stimullation", "DS_NS_Stimullation", "Log2FC_NS_Stimulation")
rotival.l <- lapply(rotival.l, function(df) setNames(df, nm= update.names))

# Remove cases where differential splicing could not be determined confidently due to low expression (labelled in DS_NS_Stimullation as "n.d")
rotival.l <- lapply(rotival.l, function(df) df[! df$DS_NS_Stimullation %in% "n.d.", ])

# Select one AS event per gene, because we will do the overlp with genes. Keep the most significant 
rotival.l <- lapply(rotival.l, function(df){
  as.data.frame( df %>% group_by(gene_symbol) %>% slice_min(FDR_NS_Stimullation, with_ties = FALSE) ) 
})

# Rename certain columns to match ours
rotival.l <- lapply(rotival.l, function(x){
  colnames(x)[ match(c("gene_symbol", "FDR_NS_Stimullation"), colnames(x)) ] <- c("gene_name", "fdr")
  return(x)
  })

# Save processed Rotival data
file.name <- "05_infectionDTU-Overlaps/01_Processed-Data/00_Rotival2019-infection-DS-parsed.rds"
saveRDS(rotival.l, file.name)
```

### 2.1.2 Overlap 

- Our data Parse 
```{r}
# Load unfiltered data
dtu.l <- readRDS("01_Processed-Data/05_Filt-1-5-0.8-0.01/01_infectionDTU-unfiltered-list.rds")
# select 
# dtu.l <- select.df.from.nested.list(dtu.l, "Intercept")
# select lineage
dtu.l <- dtu.l$lineage
# restructure to Condition > celltype
dtu.l <- list("COV.NS" = lapply(dtu.l, function(x) x[["COV.NS"]]), 
              "IAV.NS" = lapply(dtu.l, function(x) x[["IAV.NS"]]), 
              "IAV.COV" = lapply(dtu.l, function(x) x[["IAV.COV"]]))
```

- Overlap Recall *Rotival considers Differentially Spliced the |ΔPSI| > 0.05 and FDR < 0.05*
```{r}
# Read Rotival 
file.name <- "05_infectionDTU-Overlaps/01_Processed-Data/00_Rotival2019-infection-DS-parsed.rds"
rotival.l <- readRDS(file.name)

overlap.l <- lapply(dtu.l, function(condition_1){
  lapply(condition_1, function(celltype){
    lapply(rotival.l, function(condition_2){
      test.overlap.pro(m1 = celltype, fdr.col = "fdr", fdr.thres.1 = 0.05, gene.col.1 = "gene_name",
                       FC.col.1 = NA, FC.thres.1 = NA, 
                       m2 = condition_2, fdr.col.2 = "fdr", fdr.thres.2 = 0.05, gene.col.2 = "gene_name", 
                       FC.col.2 = "Delta_psi_NS_Stimullation", FC.thres.2 = 0.05) # 
    })
  })
})

# Merge
overlap.l <- lapply(overlap.l, function(condition_1) lapply(condition_1, function(celltype) plyr::ldply(celltype, .id = "Rotival")))
overlap.l <- lapply(overlap.l, function(condition_1) plyr::ldply(condition_1, .id = "celltype"))  
df = plyr::ldply(overlap.l, .id = "Aquino")
# Save
file.name <- "05_infectionDTU-Overlaps/01_Processed-Data/05_Filt-1-5-0.8-0.01/00_infectionDTU-Aquino2023-vs-Rotival2019.rds"
saveRDS(df, file.name)
```

- Read
```{r}
file.name <- "05_infectionDTU-Overlaps/01_Processed-Data/05_Filt-1-5-0.8-0.01/00_infectionDTU-Aquino2023-vs-Rotival2019.rds"
df <- readRDS( file.name )
```

- Add significance label
```{r}
require(dplyr)
# Process df to include y coordinate of label 
df$model.2.specific <- NULL # model_2 genes will appear in another plot
df.i <- reshape2::melt(df[ , c("Aquino", "Rotival", "celltype", "shared", "model.1.specific" )]) %>% group_by(celltype, Aquino,Rotival) %>% mutate(label_y = sum(value)) %>% ungroup()
# add p_value
df = merge(df.i,
      df [, c("celltype", "Aquino", "Rotival", "p_value") ], 
      by = c("celltype", "Aquino", "Rotival"))

df = add.p_value.label(df)
```

### 2.1.3 Plots

#### Odds Ratio Plot

```{r}
file.name <- "05_infectionDTU-Overlaps/01_Processed-Data/05_Filt-1-5-0.8-0.01/00_infectionDTU-Aquino2023-vs-Rotival2019.rds"
df <- readRDS( file.name )
# subset to comparisons and cell types of interest
df <- df[ df$Rotival %in% "NS.IAV" & df$celltype == "MONO" & df$Aquino %in% c("IAV.NS", "COV.NS"), ]
# add p_value_label
df = add.p_value.label(df)
# Factor
df$Condition <- factor(df$Aquino, levels = rev(names(condition.pairwise.vec.short)))
df$Condition <- plyr::revalue(df$Condition, condition.pairwise.vec.short) 
color.vec <- condition.pairwise.color.vec.short
# Plot
odds.plt <- ggplot(data = df, aes(x = odds.ratio, y = Condition, color = Condition)) + 
  geom_point(size = 1) + 
  geom_text(aes(x = odds.ratio, y = Condition, label = p_value_label), color = "black", vjust = -0.5, size = 3.5/.pt) + 
  geom_vline(aes(xintercept = 1), size = .25, linetype = "dashed", linewidth=0.25) +
  geom_errorbarh(aes(xmax = conf.int.upper, xmin = conf.int.lower), size = .5, height = .2, linewidth=0.25) + 
  scale_color_manual(guide = "none", values = color.vec) + 
  labs(title = "tit", subtitle = "sub_tit") + ylab("y_lab") + xlab("Odds Ratio") + 
  plot_theme + theme(axis.title.y = element_blank())
  # theme(panel.background = element_blank(), plot.background = element_blank(), 
  #       axis.line.x = element_line(), axis.line.y = element_line(),
  #       axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5), 
  #       axis.text.y = element_text(size = 10, angle = 0, hjust = 0.5), axis.title.y = element_blank(), 
  #       legend.position = "none")

# Save 
file.name <- "05_infectionDTU-Overlaps/02_Plots/05_Filt-1-5-0.8-0.01/02_overlap-Rotival-only-MONO-01_OR.pdf" 
ggsave(file.name, odds.plt, width = 4, height = 2, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, odds.plt, width = 4, height = 2, units = "cm", bg = "white")
```

#### BarPlot

*Display a barplot with number of Rotival2019-specific, shared, and our-study-specific*
```{r}
file.name <- "05_infectionDTU-Overlaps/01_Processed-Data/05_Filt-1-5-0.8-0.01/00_infectionDTU-Aquino2023-vs-Rotival2019.rds"
df <- readRDS( file.name )
# subset to comparisons and cell types of interest
df <- df[ df$Rotival %in% "NS.IAV" & df$celltype == "MONO" & df$Aquino %in% c("IAV.NS", "COV.NS"), ]
# add p_value_label
df = add.p_value.label(df)
# Factor
df$Condition <- factor(df$Aquino, levels = rev(names(condition.pairwise.vec.short)))
df$Condition <- plyr::revalue(df$Condition, condition.pairwise.vec.short) 
color.vec <- condition.pairwise.color.vec.short

df <- reshape2::melt( df[, c("Condition", "shared", "model.1.specific", "model.2.specific") ], 
                      value.name = "n_genes", variable.name = "model")
df$model <- plyr::revalue(df$mode, c("model.1.specific" = "Aquino 2023", "model.2.specific" = "Rotival 2019")) 
color.vec <- c("Aquino 2023" = "#4a4e69", "shared" = "#f2e9e4", "Rotival 2019" = "#c9ada7")
df$model <- factor(df$model, levels = c("Aquino 2023", "shared", "Rotival 2019"))

plt <- ggplot(data = df, aes(x = n_genes, y = Condition, fill = model)) + 
  geom_bar(stat="identity", color = "black", linewidth=0.25, position = position_stack())  +
  geom_text(aes(x = n_genes, label = n_genes), vjust=0.5, color= "black", size=3.5/.pt,  
            position = position_stack(vjust = .5 )) +
  # geom_text(aes(y = Condition, x = label_x, label = p_value_label),angle = 90, 
  #           vjust = 1.25, hjust = 0.5, size = 3) +  # p_value
  scale_fill_manual("", values = color.vec,  drop = F) +
  labs(title = "infection DTU sharing", subtitle = "Aquino infection DTU vs ROtival 2019") +
  xlab("N infection DTU Genes") + 
  plot_theme + 
  theme(axis.title.y = element_blank())
  # theme(panel.background = element_blank(), plot.background = element_blank(), 
  #       axis.line.x = element_line(), axis.line.y = element_line(),
  #       axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5), 
  #       axis.text.y = element_text(size = 10, angle = 0, hjust = 0.5), axis.title.y = element_blank()) + 
  # guides(fill = guide_legend(reverse=TRUE))

# Save
file.name <- "05_infectionDTU-Overlaps/02_Plots/05_Filt-1-5-0.8-0.01/02_overlap-Rotival-only-MONO-02_BarPlot.pdf" 
ggsave(file.name, plt, width = 5, height = 2, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, width = 5, height = 2, units = "cm", bg = "white")
```
