---
title: "01_DA"
author: "Ruben Chazarra Gil"
date: "`r Sys.Date()`"
output: html_document
---

# 0. Goal

Supplementary Fig. 15. Summary of tested genetic variants and genes in transcript usage quantitative trait loci (TU-QTL) across immune lineages and conditions. From left to right, number of tested genetic variants (single nucleotide polimorphisms; SNP); number of significant TU-QTLs, number of tested genes, and number of genes with a significant TU-QTL across immune lineages and conditions.


- Load funcitons
```{r}
source("../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../../01_Randolph2021/05_Differential-Analyses/02_DTU/00_Scripts/00_MANTA-functions.R")
source("../../00_MetaData/03_Aesthetics-repo.R")
setwd("../06_sQTL/03_DA/")
```

# Supp. Fig. 15

*Plot N of tested genes, SNPs, and N of significant genes, SNPs*

```{r}
## 1. N of tested variants
df <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-tested-df.rds")
df <- df %>% group_by(lineage, Condition) %>% summarise(n_snps = n_distinct(snpId))
df$Condition <- factor(df$Condition, levels = names(condition.vec.short) )
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)
df$lineage <- factor(df$lineage, levels = rev(names(lineage.color.vec)))

# Plot 
plt.1 <- ggplot(data = df, aes(x = n_snps, y = lineage, fill = lineage, color = lineage)) +
  geom_bar(stat= "identity", linewidth=0.25) + 
  geom_text(data = df, aes(x = n_snps, label = n_snps), color = "black",  size = 5/.pt, hjust = 0.5, position = position_stack(vjust = .5)) +
  facet_grid(Condition ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_fill_manual( values = lineage.color.vec ) +
  scale_color_manual( values = lineage.color.vec ) +
  xlab("N Tested SNPs") + 
  plot_theme + theme(axis.title.y = element_blank()) +  guides(fill = FALSE,  color = FALSE )

## 2. N of significant variants
df <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
df <- df %>% group_by(lineage, Condition) %>% summarise(n_snps = n_distinct(snpId))
df$Condition <- factor(df$Condition, levels = names(condition.vec.short) )
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)
df$lineage <- factor(df$lineage, levels = rev(names(lineage.color.vec)))

# Plot 
plt.2 <- ggplot(data = df, aes(x = n_snps, y = lineage, fill = lineage, color = lineage)) +
  geom_bar(stat= "identity", linewidth=0.25) + 
  geom_text(data = df, aes(x = n_snps, label = n_snps), color = "black",  size = 5/.pt, hjust = 0.5, position = position_stack(vjust = .5)) +
  facet_grid(Condition ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_fill_manual( values = lineage.color.vec ) +
  scale_color_manual( values = lineage.color.vec ) +
  xlab("N of TU-QTLs") + 
  plot_theme + theme(axis.title.y = element_blank()) + guides(fill = FALSE,  color = FALSE )

## 3. N Tested Genes and variants
df <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-tested-df.rds")
df <- df %>% group_by(lineage, Condition) %>% summarise(n_genes = n_distinct(geneId))
df$Condition <- factor(df$Condition, levels = names(condition.vec.short) )
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)
df$lineage <- factor(df$lineage, levels = rev(names(lineage.color.vec)))

# Plot 
plt.3 <- ggplot(data = df, aes(x = n_genes, y = lineage, fill = lineage, color = lineage)) +
  geom_bar(stat= "identity", linewidth=0.25) + 
  geom_text(data = df, aes(x = n_genes, label = n_genes), color = "black",  size = 5/.pt, hjust = 0.5, position = position_stack(vjust = .5)) +
  facet_grid(Condition ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_fill_manual( values = lineage.color.vec ) +
  scale_color_manual( values = lineage.color.vec ) +
  xlab("N Tested Genes") + 
  plot_theme + theme(axis.title.y = element_blank()) +  guides(fill = FALSE,  color = FALSE )

## 4. N significant genes
df <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
df <- df %>% group_by(lineage, Condition) %>% summarise(n_genes = n_distinct(geneId))
df$Condition <- factor(df$Condition, levels = names(condition.vec.short) )
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)
df$lineage <- factor(df$lineage, levels = rev(names(lineage.color.vec)))

# Plot 
plt.4 <- ggplot(data = df, aes(x = n_genes, y = lineage, fill = lineage, color = lineage)) +
  geom_bar(stat= "identity", linewidth=0.25) + 
  geom_text(data = df, aes(x = n_genes, label = n_genes), color = "black",  size = 5/.pt, hjust = 0.5, position = position_stack(vjust = .5)) +
  facet_grid(Condition ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_fill_manual( values = lineage.color.vec ) +
  scale_color_manual( values = lineage.color.vec ) +
  xlab("N Genes with a TU-QTL") + 
  plot_theme + theme(axis.title.y = element_blank()) +  guides(fill = FALSE,  color = FALSE )

## Merge plots
plt <- cowplot::plot_grid(
  plt.1, 
  plt.2 + theme(strip.text.y = element_blank(), strip.background = element_blank(),
                axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank()),
  plt.3 + theme(strip.text.y = element_blank(), strip.background = element_blank(),
               axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank()),
  plt.4 + theme(strip.text.y = element_blank(), strip.background = element_blank(),
                axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank()),
    nrow = 1, align = "h", axis = "tb" , rel_widths =  c(0.3, 0.23, 0.23, 0.24)
    )

file.name <- "02_Plots/03_celltype-fractions/01_06_01_Tested-and-significants-genes-and-variants.pdf"
ggsave(file.name, plt, width = 12, height = 6, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, width = 12, height = 5, units = "cm", bg = "white")
```


# Supp. Fig 17.A 

*Plot % popDTU genes with a TU-QTL when stratifying popDTU by MD*

```{r}
# TU QTLs
qtl <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
select.qtl <- c("lineage", "Condition", "geneId", "gene_name", "snpId", "best.snp", "md", "fdr")
qtl <- qtl[, select.qtl]
# Keep best, we only want to know if a popDTU has at least one TU-QTL
qtl <- qtl[qtl$snpId == qtl$best.snp, ]

# Pop DTU
dtu <- readRDS("../../03_Differential/02_DTU/02_DTU/02_popDTU/02_DA/01_Processed-Data/06_celltype-fractions/01_popDTU-filtered-df.rds")
dtu <- dtu$lineage
select.dtu <- c("celltype", "Condition", "gene_id", "gene_name", "MD", "fdr")
dtu <- dtu[, select.dtu ]
dtu$MD <- as.numeric(dtu$MD)
# Add MD Quantiles by lineage-Condition 
dtu$celltype_Condition <- paste0(dtu$celltype, "_", dtu$Condition)
dtu.l <- split(dtu, dtu$celltype_Condition)
dtu.l <- lapply(dtu.l, function(x){
  q.vec =  quantile(x$MD, probs = seq(0, 1, 0.25), include_lowest  = T)
  x$MD_quantile <-  cut(x$MD, breaks = q.vec, labels = paste0("Q", c(1:4)),  include.lowest= T)
  x
})
dtu <- data.table::rbindlist(dtu.l)

# Merge
df <- merge(dtu, qtl,
            by.x = c("celltype", "Condition", "gene_id", "gene_name"), 
            by.y = c("lineage", "Condition", "geneId", "gene_name"), 
            all.x = T, all.y = F)
# Label genes with
df$is.QTL <- ifelse(is.na(df$snpId), FALSE, TRUE)

# Stratify popDTU by MD
df <- df %>% group_by(celltype, Condition, MD_quantile) %>% summarise(n_genes_qtl = sum(is.QTL == "TRUE"),
                                                                      n_genes = n_distinct(gene_id))
df$n_genes_qtl_rel  = df$n_genes_qtl / df$n_genes

# Plot
df$bar_label <- paste0(df$n_genes_qtl, "/", df$n_genes)
df$Condition <- factor(df$Condition, levels = names(condition.vec.short) )
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)
color.vec <- setNames(lineage.df$color, nm = lineage.df$lineage)
df$celltype <- factor(df$celltype, levels = rev(names(color.vec)))
df$MD_quantile <- factor(df$MD_quantile, levels = rev(c("Q1", "Q2", "Q3", "Q4")))
alpha.vec <- c("Q1" = "Very Low", "Q2" = "Low", "Q3" = "Mid", "Q4" = "High")

# Plotce
plt <- ggplot(data = df, aes(y = celltype, x = n_genes_qtl_rel, fill = celltype)) +
  geom_bar( data = df, aes(alpha = MD_quantile), stat = "identity", linewidth=0.25, width = 0.8, position_dodge( width=0.8 )) +
  geom_text(aes(x = n_genes_qtl_rel, label = bar_label, group = MD_quantile ),
            alpha = 1, color = "black", size = 3/.pt, hjust = -0.1, vjust = 0.5, position = position_dodge( width=0.8 )) +
  facet_grid(Condition ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_fill_manual( values = lineage.color.vec ) +
  scale_alpha_manual("Effect size", values = rev(c(0.4, 0.6, 0.8, 1)), labels = alpha.vec) +
  xlab("PopDTU genes with a TU-QTL (%)") + ylab("y_lab") + 
  plot_theme + theme(axis.title.y = element_blank(), 
                     legend.position=c(0.86, 0.5)) + 
  guides(fill = FALSE, color = FALSE, alpha = guide_legend(reverse = T))
plt

# Save
file.name <- paste0("02_Plots/03_celltype-fractions/03_03_PopDTU-fraction-with-QTL-stratified-by-MD_01.pdf")
ggsave(file.name, plt, width = 5.5, height = 7, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, width = 5.5, height = 7, units = "cm", bg = "white")
```


# Supp. Fig 17.B 

*Compare effect sizes (MDs) of population DTU genes vs TU-QTLs* 

```{r}
# TU QTLs
qtl <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
# maybe we could select best SNP
select.qtl <- c("lineage", "Condition", "geneId", "gene_name", "snpId", "best.snp", "md", "fdr")
qtl <- qtl[, select.qtl]

# Pop DTU
dtu <- readRDS("../../03_Differential/02_DTU/02_DTU/02_popDTU/02_DA/01_Processed-Data/06_celltype-fractions/01_popDTU-filtered-df.rds")
dtu <- dtu$lineage
select.dtu <- c("celltype", "Condition", "gene_id", "gene_name", "MD", "fdr")
dtu <- dtu[, select.dtu ]
dtu$MD <- as.numeric(dtu$MD)

df <- merge(dtu, qtl,
             by.x = c("celltype", "Condition", "gene_id", "gene_name"), 
             by.y = c("lineage", "Condition", "geneId", "gene_name"), 
             all.x = T, all.y = F)

# some sQTLseekeR2 md values are NA, remove
df <- df[!is.na(df$md), ]

# Select best SNP 
# df <- df[df$snpId == df$best.snp,  ]

# Plot 
df$Condition <- factor(df$Condition, levels = names(condition.vec.short) )
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)
color.vec <- setNames(lineage.df$color, nm = lineage.df$lineage)
df$celltype <- factor(df$celltype, levels = lineage.df$lineage)

plt <- ggplot(data = df, aes(x = md, y = MD, color = celltype )) +
  # xlim(0, 0.5) + ylim(0, 0.5) + 
  geom_point(shape = 20, size = 0.05, alpha = 1)  + 
  geom_smooth(aes(fill = celltype), method = "lm", size = 0.5) +  # Add trend line
  ggpubr::stat_cor(aes(color = celltype), method = "spearman", # for all the points
                   label.x = 0, label.y = 0.2, size = 1.5) +
  facet_grid(  Condition ~ celltype, switch = "y") +
  scale_color_manual( "", values = lineage.color.vec ) +
  scale_fill_manual( "", values = lineage.color.vec ) +
  # scale_x_continuous(breaks = c(0, 0.2, 0.4)) + 
  # scale_y_continuous(breaks = c(0, 0.2, 0.4)) + 
  labs(title = "tit", subtitle = "sub_tit") + xlab("Population DTU gene MD value") + ylab("TU-QTL MD value") + 
  plot_theme + 
  guides( fill = FALSE, color = FALSE)
plt
file.name <- paste0("02_Plots/03_celltype-fractions/03_02_MD-corr-TU-QTL-PopDTU_01.pdf")
ggsave(file.name, plt, width = 11, height = 7, units = "cm")
file.name <- gsub(".pdf", ".svg",  file.name)
ggsave(file.name, plt, width = 11, height = 7, units = "cm", bg = "white")
```


# Supp Fig. 18. A

*Lineage Sharing*
```{r}
file.name <- "../../03_Differential/02_DTU/02_DTU/02_popDTU/02_DA/01_Processed-Data/06_celltype-fractions/01_popDTU-unfiltered-list.rds"
dtu.l <- readRDS(file.name)
dtu.l <- dtu.l$lineage

dtu.l <- list(
  NS = lapply(dtu.l, function(x) x[["NS"]]), 
  IAV = lapply(dtu.l, function(x) x[["IAV"]]), 
  COV = lapply(dtu.l, function(x) x[["COV"]])
)

dtu.l <- lapply(dtu.l, function(c){
  vec = Reduce(intersect, lapply(c, function(x) x$gene_name))
  lapply(c, function(x) x[x$gene_name %in% vec, ])
  })
dtu.l <- lapply(dtu.l, function(c) plyr::ldply(c, .id = "lineage"))
dtu <- plyr::ldply(dtu.l, .id = "Condition")
dtu <- dtu[dtu$fdr < 0.05, ] # keep DTU

## Merge with QTLs
qtl <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
qtl$is.QTL <- "TRUE"
qtl <- unique(qtl[, c("lineage", "Condition", "gene_name", "is.QTL")])
dtu <- merge(dtu, qtl, by = c("lineage", "Condition", "gene_name"), all.x = T, all.y = F)
dtu[is.na(dtu$is.QTL), "is.QTL"] <- "FALSE"
dtu$is.QTL <- plyr::revalue(dtu$is.QTL, c("TRUE" = "popDTU with TU-QTL", "FALSE" = "popDTU without TU-QTL"))

# Count sharing
df <- dtu %>% group_by(Condition, is.QTL, gene_name) %>% summarise(n_lineages = n_distinct(lineage))
df <- df %>% group_by(Condition, is.QTL, n_lineages) %>% summarise(n_genes = n_distinct(gene_name))
df <- df %>% group_by(Condition, is.QTL) %>% mutate(n_genes_rel = n_genes / sum(n_genes) )

df$n_lineages <- factor(df$n_lineages, levels = c(1:5))
df$Condition <- factor(df$Condition, levels = rev(names(condition.vec.short)))
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)

plt <- ggplot(df, aes(x = n_genes_rel, y = Condition, fill = n_lineages)) + 
  geom_bar(stat = "identity", color = "black", linewidth = 0.25, position = position_stack(reverse = T)) + 
   geom_text(data = df[df$n_lineages %in% c("1", "2", "3"), ],  aes(label = round(n_genes_rel, 2)), 
             position = position_stack(vjust = 0.5, reverse = T), color="black", size=3.5/.pt, angle = 90) + 
  scale_fill_brewer("N lineages", palette = "Blues") + 
  xlab("% of population DTU Genes") + 
  facet_grid(  ~is.QTL, switch = "y") +
  plot_theme + 
  theme(axis.title.y = element_blank(), 
        legend.box.spacing = unit(0.2, "cm"), legend.key.size = unit(0.2, "cm"),
        legend.title = element_text(size = 5, margin = margin(b = 0, unit = "cm"))
        )+ 
  guides(fill = guide_legend(byrow = FALSE, override.aes = list(color = "white", linewidth = 0.15)))

file.name <- "02_Plots/03_celltype-fractions/03_04_02_popDTU-lineage-sharing-with-or-without-QTL.pdf"
ggsave(file.name, plt, width = 10.5, height =5, units = "cm")
```

# Supp Fig. 18. B
*Condition sharing*
```{r}
file.name <- "../../03_Differential/02_DTU/02_DTU/02_popDTU/02_DA/01_Processed-Data/06_celltype-fractions/01_popDTU-unfiltered-list.rds"
dtu.l <- readRDS(file.name)
dtu.l <- dtu.l$lineage

dtu.l <- lapply(dtu.l, function(ct){
  vec = Reduce(intersect, lapply(ct, function(x) x$gene_name))
  lapply(ct, function(x) x[x$gene_name %in% vec, ])
  })
dtu.l <- lapply(dtu.l, function(ct) plyr::ldply(ct, .id = "Condition"))
dtu <- plyr::ldply(dtu.l, .id = "lineage")
dtu <- dtu[dtu$fdr < 0.05, ] # keep DTU

## Merge with QTLs
qtl <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
qtl$is.QTL <- "TRUE"
qtl <- unique(qtl[, c("lineage", "Condition", "gene_name", "is.QTL")])
dtu <- merge(dtu, qtl, by = c("lineage", "Condition", "gene_name"), all.x = T, all.y = F)
dtu[is.na(dtu$is.QTL), "is.QTL"] <- "FALSE"
dtu$is.QTL <- plyr::revalue(dtu$is.QTL, c("TRUE" = "popDTU with TU-QTL", "FALSE" = "popDTU without TU-QTL"))

df <- dtu %>% group_by(lineage, is.QTL, gene_name) %>% summarise(n_conditions = n_distinct(Condition), Condition = paste(Condition, collapse = ","))
df <- df %>% group_by(lineage, is.QTL, Condition) %>% summarise(n_genes = n_distinct(gene_name))
df <- df %>% group_by(lineage, is.QTL) %>% mutate(n_genes_rel = n_genes /sum(n_genes))

df$Condition <- plyr::revalue(df$Condition, c("NS" = "Baseline", 
                                              "COV,NS" = "Baseline,COV", "IAV,NS" = "Baseline,IAV", 
                                              "COV,IAV" = "IAV,COV", "COV,IAV,NS" = "All"))

df$Condition <- factor(df$Condition, levels = c("Baseline", "IAV", "COV","Baseline,IAV","Baseline,COV", 
                                                "IAV,COV", "All"))
df$lineage <- factor(df$lineage, levels = rev(lineage.df$lineage))
color.vec <- c("Baseline" = "#edf2f4", "IAV" = "#75A9D1", "COV" = "#c1121f", 
               "Baseline,COV" = "#780000", "Baseline,IAV" = "#003049", 
               "IAV,COV" = "#ffb41f", "All" = "#fdf0d5")

plt <- ggplot(df, aes(y = lineage, x = n_genes_rel, fill = Condition)) + 
  geom_bar(width=0.8, stat = "identity", #color = "white", 
           linewidth = 0.25, position = position_stack(reverse = T)) + 
  geom_text(aes(label = round(n_genes_rel, 2)), position = position_stack(vjust = 0.5, reverse = T),
            color="black", alpha = 1, size=3/.pt, angle = 90) +
  scale_fill_manual(values = color.vec ) +
  xlab("% of Population DTU Genes") + 
  facet_grid(   ~is.QTL , scales = "free_x", switch = "y") +
  plot_theme  + theme(axis.title.y = element_blank(), axis.line.y = element_blank())
plt

file.name <- "02_Plots/03_celltype-fractions/03_04_03_popDTU-Condition-sharing-with-or-without-QTL.pdf"
ggsave(file.name, plt, width = 10.5, height = 5, units = "cm")
```





# Supp. Fig. 19. A

*Lineage sharing stratifying by Condition*
*Consider intersection of tested genes*
```{r}
file.name <- "01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-tested-list.rds"
l <- readRDS(file.name)

# Add geneId_snpId 
l <- lapply(l, function(c) lapply(c, function(x){ x$geneId_snpId <- paste0(x$geneId, "_", x$snpId); x}))

# Restructure 
l <- list(
  NS = lapply(l, function(x) x[["NS"]]), 
  IAV = lapply(l, function(x) x[["IAV"]]), 
  COV = lapply(l, function(x) x[["COV"]])
)
# Keep QTLs tested in all lineages
l <- lapply(l, function(c){
  vec = Reduce(intersect, lapply(c, function(x) x$geneId_snpId))
  lapply(c, function(x) x[x$geneId_snpId %in% vec, ])
  })
# Re-merge
l <- lapply(l, function(c) plyr::ldply(c, .id = "lineage"))
qtl <- plyr::ldply(l, .id = "Condition")

# Add fdrs from filtered set 
sig <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
sig <- sig[, c("Condition", "lineage", "geneId", "snpId", "fdr")]
sig$geneId_snpId <- paste0(sig$geneId, "_", sig$snpId)
qtl <- merge(qtl, sig, all.x = F, all.y = F) # By this we keep the significant ones from the set of common set of geneId_snpId tested across lineages whitin condition

df <- qtl %>% group_by(Condition, geneId_snpId) %>% summarise(n_lineages = n_distinct(lineage))
df <- df %>% group_by(Condition, n_lineages) %>% summarise(n_spns = n_distinct(geneId_snpId))
df <- df %>% group_by(Condition) %>% mutate(n_spns_rel = n_spns / sum(n_spns) )

df$n_lineages <- factor(df$n_lineages, levels = c(1:5))
df$Condition <- factor(df$Condition, levels = rev(names(condition.vec.short)))
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)

plt <- ggplot(df, aes(y = Condition, x = n_spns_rel, fill = n_lineages)) + 
  geom_bar(stat = "identity", color = "black", linewidth = 0.25, position = position_stack(reverse = T)) + 
   geom_text(data = df, aes(label = round(n_spns_rel, 2)), position = position_stack(vjust = 0.5, reverse = T),
             color="black", size=3.5/.pt, angle = 90) + 
  scale_fill_brewer("N lineages", palette = "Blues") + 
  xlab("% TU-QTLs") + 
  # facet_grid( Condition ~., scales = "free_x", switch = "y") +
  plot_theme + 
  theme(axis.title.y = element_blank(), axis.line.y = element_blank(),
        #axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 3.5, margin = margin(b = 0)), 
        legend.box.spacing = unit(0.2, "cm"),
        legend.key.size = unit(0.2, "cm"),
        legend.title = element_text(size = 5, margin = margin(b = 0, unit = "cm"))
        )+ 
  guides(fill = guide_legend(byrow = FALSE, override.aes = list(color = "white", linewidth = 0.15)))

file.name <- "02_Plots/03_celltype-fractions/02_03_01_TU-QTL-lineage-sharing-02_by-Condition-01_horizontal.pdf"
ggsave(file.name, plt, width = 5, height = 5, units = "cm")
```


# Supp Fig. 19. B

*Condition sharing*

*Assessing  condition sharing within each immune lineage is "more fair" than lineage sharing across condition because of sample size. However, we subset to intersection of tested genes to make a more robust assessment*

```{r}
file.name <- "01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-tested-list.rds"
l <- readRDS(file.name)

# Add geneId_snpId 
l <- lapply(l, function(c) lapply(c, function(x){ x$geneId_snpId <- paste0(x$geneId, "_", x$snpId); x}))

# Keep QTLs tested in all lineages
l <- lapply(l, function(c){
  vec = Reduce(intersect, lapply(c, function(x) x$geneId_snpId))
  lapply(c, function(x) x[x$geneId_snpId %in% vec, ])
  })

# Re-merge
l <- lapply(l, function(ct) plyr::ldply(ct, .id = "Condition"))
qtl <- plyr::ldply(l, .id = "lineage")

# Add fdrs from filtered set 
sig <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
sig <- sig[, c("Condition", "lineage", "geneId", "snpId", "fdr")]
sig$geneId_snpId <- paste0(sig$geneId, "_", sig$snpId)
qtl <- merge(qtl, sig, all.x = F, all.y = F)

# Count sharing
df <- qtl %>% group_by(lineage, geneId_snpId) %>% summarise(n_conditions = n_distinct(Condition), Condition = paste(Condition, collapse = ","))
df <- df %>% group_by(lineage, Condition) %>% summarise(n_snps = n_distinct(geneId_snpId))
df <- df %>% group_by(lineage) %>% mutate(n_snps_rel = n_snps /sum(n_snps))

df$Condition <- plyr::revalue(df$Condition, c("NS" = "Baseline", 
                                              "COV,NS" = "Baseline,COV", "IAV,NS" = "Baseline,IAV", 
                                              "COV,IAV" = "IAV,COV", "COV,IAV,NS" = "All"))

df$Condition <- factor(df$Condition, levels = c("Baseline", "IAV", "COV","Baseline,IAV","Baseline,COV", 
                                                "IAV,COV", "All"))
df$lineage <- factor(df$lineage, levels = rev(lineage.df$lineage))
color.vec <- c("Baseline" = "#edf2f4", "IAV" = "#75A9D1", "COV" = "#c1121f", 
               "Baseline,COV" = "#780000", "Baseline,IAV" = "#003049", 
               "IAV,COV" = "#ffb41f", "All" = "#fdf0d5")

plt <- ggplot(df, aes(y = lineage, x = n_snps_rel, fill = Condition)) + 
  geom_bar(width=0.8, stat = "identity", #color = "white", 
           linewidth = 0.25, position = position_stack(reverse = T)) + 
  geom_text(aes(label = round(n_snps_rel, 2)), position = position_stack(vjust = 0.5, reverse = T),
            color="black", alpha = 1, size=3/.pt, angle = 90) +
  scale_fill_manual(values = color.vec ) +
  xlab("% TU-QTLs") + 
  plot_theme  + theme(axis.title.y = element_blank(), axis.line.y = element_blank())
plt

file.name <- "02_Plots/03_celltype-fractions/02_03_02_TU-QTL-Condition-sharing-02_by-Lineage-01_horizontal.pdf"
ggsave(file.name, plt, width = 5.5, height = 5, units = "cm")
```



# Supp. Fig. 19. C

- For plots with TU Gene or QTL *Change params*

- Consider genes tested in all conditions
```{r}
l <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-tested-list.rds")
l <- unlist(l, recursive = F)
gene.l <- lapply(l, function(x) unique(x$geneId))
gene.vec <- Reduce(intersect, gene.l)
```

```{r}
t <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
# Subset to genes tested across all lineage-conditions
t <- t[t$geneId %in% gene.vec, ]
df <- t %>% distinct(lineage, Condition, gene_name) # for gene_names
# df <- t %>% distinct(lineage, Condition, snpId) # for QTLs
l <- split(df, df$lineage)
l <- lapply(l, function(x) split(x, x$Condition))
gene.l <- lapply(l, function(x) lapply(x, function(y) y$gene_name)) # for gene_names
#gene.l <- lapply(l, function(x) lapply(x, function(y) y$snpId)) # for QTLs
# unlist one level
gene.l <- unlist(gene.l, recursive = F)
#Pre-process for upset # prepare.ComplexUpset() is preloaded
df <- prepare.ComplexUpset(gene.l)
```

- ComplexUpset Plot 
```{r}
# naming
colnames(df) <-  gsub("(.*)\\.(.*\\..*)", "\\1_\\2", colnames(df))
name.df <- data.frame(ct_condition = colnames(df))
name.df$lineage <- gsub("\\..*", "", name.df$ct_condition)
name.df$lineage <- gsub("_", ".", name.df$lineage)
name.df$Condition <- gsub(".*\\.", "", name.df$ct_condition)
name.df <- merge(name.df,  lineage.df[, c("lineage", "color")], by = "lineage", all = T)
name.df$Condition_2 <- plyr::revalue(name.df$Condition, condition.vec.short)
name.df$lineage_Condition_2 <- paste0(name.df$lineage, "  ", name.df$Condition_2)
# subset name.df
name.df <- name.df[name.df$ct_condition %in% colnames(df), ]
name.vec <- setNames(name.df$lineage_Condition_2, nm = name.df$ct_condition)
color.vec <- setNames(name.df$color, nm = name.df$ct_condition)

min_set_size = 1
font_size = 5
# Plot
plt <- ComplexUpset::upset(df, names(df), 
                           min_size = min_set_size,
                           set_sizes = FALSE, 
                           labeller = ggplot2::as_labeller(name.vec), 
                           name = NULL, 
                           height_ratio=0.8,
                           matrix=( intersection_matrix(geom=geom_point(size=0.75))), 
                    base_annotations = list(
                      # Barplot
                      'N of Genes with a TU-QTL' = intersection_size( fill = "#778da9",
                                                            text = list(size = 3/.pt, angle=90, vjust = 0.5, hjust = -0.2),
                                                            text_colors = c(on_background = "black",
                                                                            on_bar = "black")) +
                                      theme( panel.grid = element_blank(), axis.line = element_line(), 
                                             axis.title.y = element_text(size = font_size, vjust = -15),
                                             axis.text.y  = element_text(size = font_size))
                      ),
                    themes=upset_modify_themes(
                      list(
                        'intersection_size'=theme(text=element_text(size=font_size),
                                                  plot.margin = margin(0, 0, 0, 0, unit = "cm")),
                        'intersections_matrix'=theme(text=element_text(size=font_size),
                                                     plot.margin = margin(0, 0, 0, 0, unit = "cm"))
                        )), 
queries=list(upset_query(set= "T_CD4.NS", color = "#355070", fill="#355070"), # Not finding a systematic way of edditing colors !
             upset_query(set= "T_CD4.IAV", color = "#355070", fill="#355070"), 
             upset_query(set= "T_CD4.COV", color = "#355070", fill="#355070"), 
             upset_query(set= "T_CD8.NS", color = "#b56576", fill="#b56576"), 
             upset_query(set= "T_CD8.IAV", color = "#b56576", fill="#b56576"),
             upset_query(set= "T_CD8.COV", color = "#b56576", fill="#b56576"),
             upset_query(set= "MONO.NS", color = "#6d597a", fill="#6d597a"),
             upset_query(set= "MONO.IAV", color = "#6d597a", fill="#6d597a"),
             upset_query(set= "MONO.COV", color = "#6d597a", fill="#6d597a"),
             upset_query(set= "NK.NS", color = "#eaac8b", fill="#eaac8b"),
             upset_query(set= "NK.IAV", color = "#eaac8b", fill="#eaac8b"),
             upset_query(set= "NK.COV", color = "#eaac8b", fill="#eaac8b"), 
             upset_query(set= "B.NS", color = "#e56b6f", fill="#e56b6f"),
             upset_query(set= "B.IAV", color = "#e56b6f", fill="#e56b6f"), 
             upset_query(set= "B.COV", color = "#e56b6f",fill="#e56b6f")
             )
)

plt
file.name <- paste0("02_Plots/03_celltype-fractions/02_01_TU-Gene-ComplexUpset-01_lineage-min-set-", min_set_size, "-intersect-tested-genes.pdf") # for TU Gene
# file.name <- paste0("02_Plots/03_celltype-fractions/02_02_TU-QTL-ComplexUpset-01_lineage-min-set-", min_set_size, "-intersect-tested-genes.pdf") # for TU QTL
ggsave(file.name, plt, width = 18, height = 6, units = "cm")
# file.name <- gsub(".pdf", ".svg", file.name)
# ggsave(file.name, plt, width = 12, height = 5, units = "cm", bg = "white")
```

### Get N sharing
*Fractions of gene-SNP pairs shared*
```{r}
# 1. Sharing across lineage_Condition

l <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-tested-list.rds") # Consider sets tested in all conditions
l <- unlist(l, recursive=F)
l <- lapply(l, function(x) paste0(x$geneId, "_", x$snpId))
vec <- Reduce(intersect, l)


df <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
df$geneId_snpId <- paste0(df$geneId, "_", df$snpId)
df$lineage_Condition <- paste0(df$lineage, "_", df$Condition)
df <- df %>% dplyr::select(geneId_snpId, lineage_Condition)
df <- df[df$geneId_snpId %in% vec, ] # subset to variants tested in all conditions
df <- df %>% group_by(geneId_snpId) %>% summarise(n_conditions = n_distinct(lineage_Condition))
df <- df %>%  group_by(n_conditions) %>% summarise(n_qtls = n_distinct(geneId_snpId))
df$n_qtls_rel <- df$n_qtls / sum(df$n_qtls)
df.1 <- df

# 2.Condition sharing across lineages
l <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-tested-list.rds") # Consider sets tested in all conditions
l <- lapply(l, function(x)lapply(x, function(y) paste0(y$geneId, "_", y$snpId) ))
l <- lapply(l, function(x) Reduce(intersect, x))

sig.l <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-list.rds")
sig.l <- lapply(sig.l, function(x)lapply(x, function(y) paste0(y$geneId, "_", y$snpId) ))
cts <- names(sig.l)
sig.l <- lapply(cts, function(ct) lapply(sig.l[[ct]], function(y) data.frame( geneId_snpId = y[y %in% l[[ct]]]) )) # subset to common variants across conditions
names(sig.l) <- cts
sig.l <- lapply(sig.l, function(ct) plyr::ldply(ct, .id ="Condition"))
df <- plyr::ldply(sig.l, .id = "lineage")
df$lineage_Condition <- paste0(df$lineage, "_", df$Condition)
df <- df %>% group_by(lineage, geneId_snpId) %>% summarise(n_conditions = n_distinct(Condition))
df <- df %>% group_by(lineage, n_conditions) %>% summarise(n_qtls = n_distinct(geneId_snpId))
df <- df %>% group_by(lineage) %>% mutate(n_qtls_rel = n_qtls / sum(n_qtls))
df[df$n_conditions == 3, ]
df.2 <- df

# 3. Lineage sharing across Conditions
l <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-tested-list.rds") # Consider sets tested in all conditions
l <- list("NS" = lapply(l, function(x) x[["NS"]]), "IAV" = lapply(l, function(x) x[["IAV"]]), "COV" = lapply(l, function(x) x[["COV"]]))
l <- lapply(l, function(x)lapply(x, function(y) paste0(y$geneId, "_", y$snpId) ))
l <- lapply(l, function(x) Reduce(intersect, x))

sig.l <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-list.rds")
sig.l <- list("NS" = lapply(sig.l, function(x) x[["NS"]]), "IAV" = lapply(sig.l, function(x) x[["IAV"]]), "COV" = lapply(sig.l, function(x) x[["COV"]]))
sig.l <- lapply(sig.l, function(x)lapply(x, function(y) paste0(y$geneId, "_", y$snpId) ))

conditions <- names(sig.l)
sig.l <- lapply(conditions, function(c) lapply(sig.l[[c]], function(y) data.frame( geneId_snpId = y[y %in% l[[c]]]) )) # subset to common variants across conditions
names(sig.l) <- conditions
sig.l <- lapply(sig.l, function(ct) plyr::ldply(ct, .id ="lineage"))
df <- plyr::ldply(sig.l, .id = "Condition")
df <- df %>% group_by(Condition, geneId_snpId) %>% summarise(n_lineages = n_distinct(lineage))
df <- df %>% group_by(Condition, n_lineages) %>% summarise(n_qtls = n_distinct(geneId_snpId))
df <- df %>% group_by(Condition) %>% mutate(n_qtls_rel = n_qtls / sum(n_qtls))
df[df$n_lineages == 5, ]
df.3 <- df


l <- list("lineage_Condition_sharing" = df.1, "Condition_sharing_across_lineage" = df.2, "lineage_sharing_across_Condtion" = df.3)
saveRDS( l, "01_Processed-Data/03_celltype-fractions/01_02_TU-QTL-sharing-numbers.rds" )
```

# 2.3 KEEP Get N of lineage specific TU-QTLs

```{r}
qtls.all <- read.table("01_Processed-Data/03_celltype-fractions/results_all_tested.tsv", sep = "\t", header = T)
colnames(qtls.all) <- plyr::revalue(colnames(qtls.all), c("cell.lineage" = "lineage", "condition" = "Condition"))
qtls.all$Condition <- plyr::revalue(qtls.all$Condition, c("Baseline" = "NS", "Influenza A Virus" = "IAV", "SARS-CoV-2" = "COV"))
# get list of snpIds tested in all lineage-Condtions
qtls.all <- qtls.all %>% filter(lineage %in% c("T.CD4", "T.CD8", "MONO")) %>% distinct(lineage, Condition, gene_name, snpId)

# Select those SNPs common to all lineage-Conditions
l <- split(qtls.all, qtls.all$lineage)
l <- lapply(l, function(x) split(x, x$Condition))
l <- unlist(l, recursive = F)
snp.l <- lapply(l, function(x) x$snpId)
common.snps <- Reduce(intersect, snp.l)
length(common.snps)

# N of lineage specific
qtls.sig <- read.table("01_Processed-Data/03_celltype-fractions/results_all_significant.tsv", sep = "\t", header = T)
qtls.sig <- qtls.sig %>% filter(lineage %in% c("T.CD4", "T.CD8", "MONO")) %>% distinct(lineage, Condition, gene_name, snpId)
qtls.sig <- qtls.sig[qtls.sig$snpId %in% common.snps, ]
qtls.sig <- qtls.sig %>% group_by(Condition, snpId) %>% summarise(n_lineage = n_distinct(lineage), 
                                                                  lineages = paste0(unique(lineage), collapse=","))

qtls.sig <- qtls.sig %>% group_by( Condition, n_lineage, lineages ) %>% summarise(n_qtls = n_distinct(snpId))
qtls.sig <- qtls.sig %>% group_by(Condition) %>% mutate(n_qtls_rel = n_qtls/ sum(n_qtls))
# qtls.sig <- qtls.sig %>% group_by(Condition, n_lineage) %>% summarise(n_qtls = sum(n_qtls), n_qtls_rel= sum(n_qtls_rel))
qtls.sig

# N of condition specific
qtls.sig <- read.table("01_Processed-Data/03_celltype-fractions/results_all_significant.tsv", sep = "\t", header = T)
qtls.sig <- qtls.sig %>% filter(lineage %in% c("T.CD4", "T.CD8", "MONO")) %>% distinct(lineage, Condition, gene_name, snpId)
qtls.sig <- qtls.sig[qtls.sig$snpId %in% common.snps, ]
qtls.sig <- qtls.sig %>% group_by(lineage, snpId) %>% summarise(n_conditions = n_distinct(Condition), 
                                                                conditions = paste0(unique(Condition), collapse=","))

qtls.sig <- qtls.sig %>% group_by( lineage, n_conditions, conditions ) %>% summarise(n_qtls = n_distinct(snpId))
qtls.sig <- qtls.sig %>% group_by(lineage) %>% mutate(n_qtls_rel = n_qtls/ sum(n_qtls))
# qtls.sig <- qtls.sig %>% group_by(lineage, n_conditions) %>% summarise(n_qtls = sum(n_qtls), n_qtls_rel= sum(n_qtls_rel))
qtls.sig
```



