---
title: "01_DA"
author: "Ruben Chazarra Gil"
date: "`r Sys.Date()`"
output: html_document
---

# 0. Goal

*Supplementary Fig. 19.* Sharing of transcript usage quantitative trait loci (TU-QTLs) across immune lineages and conditions. 
- a) Proportion of TU-QTLs shared across immune lineages for each condition. 
- b) Proportion of TU-QTLs detected in each condition combination within each immune lineage. 
- c) UpSet plot showing the overlap of TU-QTLs across immune lineages and conditions (baseline, influenza A virus (IAV), and SARS-CoV-2 (COV)) for sets >4. A common set of TU-QTLs tested across immune lineages and conditions is considered across the different plots


- Load functions
```{r}
source("../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../../01_Randolph2021/05_Differential-Analyses/02_DTU/00_Scripts/00_MANTA-functions.R")
source("../../00_MetaData/03_Aesthetics-repo.R")
setwd("../06_sQTL/03_DA/")
```

# Supp. Fig. 19. A

*Supplementary Fig. 19.*  a) Proportion of TU-QTLs shared across immune lineages for each condition. 
```{r}
file.name <- "01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-tested-list.rds"
l <- readRDS(file.name)

# Add geneId_snpId 
l <- lapply(l, function(c) lapply(c, function(x){ x$geneId_snpId <- paste0(x$geneId, "_", x$snpId); x}))

# Restructure 
l <- list(
  NS = lapply(l, function(x) x[["NS"]]), 
  IAV = lapply(l, function(x) x[["IAV"]]), 
  COV = lapply(l, function(x) x[["COV"]])
)
# Keep QTLs tested in all lineages
l <- lapply(l, function(c){
  vec = Reduce(intersect, lapply(c, function(x) x$geneId_snpId))
  lapply(c, function(x) x[x$geneId_snpId %in% vec, ])
  })
# Re-merge
l <- lapply(l, function(c) plyr::ldply(c, .id = "lineage"))
qtl <- plyr::ldply(l, .id = "Condition")

# Add fdrs from filtered set 
sig <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
sig <- sig[, c("Condition", "lineage", "geneId", "snpId", "fdr")]
sig$geneId_snpId <- paste0(sig$geneId, "_", sig$snpId)
qtl <- merge(qtl, sig, all.x = F, all.y = F) # By this we keep the significant ones from the set of common set of geneId_snpId tested across lineages whitin condition

df <- qtl %>% group_by(Condition, geneId_snpId) %>% summarise(n_lineages = n_distinct(lineage))
df <- df %>% group_by(Condition, n_lineages) %>% summarise(n_spns = n_distinct(geneId_snpId))
df <- df %>% group_by(Condition) %>% mutate(n_spns_rel = n_spns / sum(n_spns) )

df$n_lineages <- factor(df$n_lineages, levels = c(1:5))
df$Condition <- factor(df$Condition, levels = rev(names(condition.vec.short)))
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)

plt <- ggplot(df, aes(y = Condition, x = n_spns_rel, fill = n_lineages)) + 
  geom_bar(stat = "identity", color = "black", linewidth = 0.25, position = position_stack(reverse = T)) + 
   geom_text(data = df, aes(label = round(n_spns_rel, 2)), position = position_stack(vjust = 0.5, reverse = T),
             color="black", size=3.5/.pt, angle = 90) + 
  scale_fill_brewer("N lineages", palette = "Blues") + 
  xlab("% TU-QTLs") + 
  # facet_grid( Condition ~., scales = "free_x", switch = "y") +
  plot_theme + 
  theme(axis.title.y = element_blank(), axis.line.y = element_blank(),
        #axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 3.5, margin = margin(b = 0)), 
        legend.box.spacing = unit(0.2, "cm"),
        legend.key.size = unit(0.2, "cm"),
        legend.title = element_text(size = 5, margin = margin(b = 0, unit = "cm"))
        )+ 
  guides(fill = guide_legend(byrow = FALSE, override.aes = list(color = "white", linewidth = 0.15)))

file.name <- "02_Plots/03_celltype-fractions/02_03_01_TU-QTL-lineage-sharing-02_by-Condition-01_horizontal.pdf"
ggsave(file.name, plt, width = 5, height = 5, units = "cm")
```


# Supp Fig. 19. B

*Condition sharing*

*Supplementary Fig. 19.*  b) Proportion of TU-QTLs detected in each condition combination within each immune lineage. 

```{r}
file.name <- "01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-tested-list.rds"
l <- readRDS(file.name)

# Add geneId_snpId 
l <- lapply(l, function(c) lapply(c, function(x){ x$geneId_snpId <- paste0(x$geneId, "_", x$snpId); x}))

# Keep QTLs tested in all lineages
l <- lapply(l, function(c){
  vec = Reduce(intersect, lapply(c, function(x) x$geneId_snpId))
  lapply(c, function(x) x[x$geneId_snpId %in% vec, ])
  })

# Re-merge
l <- lapply(l, function(ct) plyr::ldply(ct, .id = "Condition"))
qtl <- plyr::ldply(l, .id = "lineage")

# Add fdrs from filtered set 
sig <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
sig <- sig[, c("Condition", "lineage", "geneId", "snpId", "fdr")]
sig$geneId_snpId <- paste0(sig$geneId, "_", sig$snpId)
qtl <- merge(qtl, sig, all.x = F, all.y = F)

# Count sharing
df <- qtl %>% group_by(lineage, geneId_snpId) %>% summarise(n_conditions = n_distinct(Condition), Condition = paste(Condition, collapse = ","))
df <- df %>% group_by(lineage, Condition) %>% summarise(n_snps = n_distinct(geneId_snpId))
df <- df %>% group_by(lineage) %>% mutate(n_snps_rel = n_snps /sum(n_snps))

df$Condition <- plyr::revalue(df$Condition, c("NS" = "Baseline", 
                                              "COV,NS" = "Baseline,COV", "IAV,NS" = "Baseline,IAV", 
                                              "COV,IAV" = "IAV,COV", "COV,IAV,NS" = "All"))

df$Condition <- factor(df$Condition, levels = c("Baseline", "IAV", "COV","Baseline,IAV","Baseline,COV", 
                                                "IAV,COV", "All"))
df$lineage <- factor(df$lineage, levels = rev(lineage.df$lineage))
color.vec <- c("Baseline" = "#edf2f4", "IAV" = "#75A9D1", "COV" = "#c1121f", 
               "Baseline,COV" = "#780000", "Baseline,IAV" = "#003049", 
               "IAV,COV" = "#ffb41f", "All" = "#fdf0d5")

plt <- ggplot(df, aes(y = lineage, x = n_snps_rel, fill = Condition)) + 
  geom_bar(width=0.8, stat = "identity", #color = "white", 
           linewidth = 0.25, position = position_stack(reverse = T)) + 
  geom_text(aes(label = round(n_snps_rel, 2)), position = position_stack(vjust = 0.5, reverse = T),
            color="black", alpha = 1, size=3/.pt, angle = 90) +
  scale_fill_manual(values = color.vec ) +
  xlab("% TU-QTLs") + 
  plot_theme  + theme(axis.title.y = element_blank(), axis.line.y = element_blank())
plt

file.name <- "02_Plots/03_celltype-fractions/02_03_02_TU-QTL-Condition-sharing-02_by-Lineage-01_horizontal.pdf"
ggsave(file.name, plt, width = 5.5, height = 5, units = "cm")
```



# Supp. Fig. 19. C

*Supplementary Fig. 19.*  c) UpSet plot showing the overlap of TU-QTLs across immune lineages and conditions (baseline, influenza A virus (IAV), and SARS-CoV-2 (COV)) for sets >4. A common set of TU-QTLs tested across immune lineages and conditions is considered across the different plots
- Consider genes tested in all conditions

```{r}
l <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-tested-list.rds")
l <- unlist(l, recursive = F)
gene.l <- lapply(l, function(x) unique(x$geneId))
gene.vec <- Reduce(intersect, gene.l)
```

```{r}
t <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
# Subset to genes tested across all lineage-conditions
t <- t[t$geneId %in% gene.vec, ]
df <- t %>% distinct(lineage, Condition, gene_name) # for gene_names
# df <- t %>% distinct(lineage, Condition, snpId) # for QTLs
l <- split(df, df$lineage)
l <- lapply(l, function(x) split(x, x$Condition))
gene.l <- lapply(l, function(x) lapply(x, function(y) y$gene_name)) # for gene_names
#gene.l <- lapply(l, function(x) lapply(x, function(y) y$snpId)) # for QTLs
# unlist one level
gene.l <- unlist(gene.l, recursive = F)
#Pre-process for upset # prepare.ComplexUpset() is preloaded
df <- prepare.ComplexUpset(gene.l)
```

- ComplexUpset Plot 
```{r}
# naming
colnames(df) <-  gsub("(.*)\\.(.*\\..*)", "\\1_\\2", colnames(df))
name.df <- data.frame(ct_condition = colnames(df))
name.df$lineage <- gsub("\\..*", "", name.df$ct_condition)
name.df$lineage <- gsub("_", ".", name.df$lineage)
name.df$Condition <- gsub(".*\\.", "", name.df$ct_condition)
name.df <- merge(name.df,  lineage.df[, c("lineage", "color")], by = "lineage", all = T)
name.df$Condition_2 <- plyr::revalue(name.df$Condition, condition.vec.short)
name.df$lineage_Condition_2 <- paste0(name.df$lineage, "  ", name.df$Condition_2)
# subset name.df
name.df <- name.df[name.df$ct_condition %in% colnames(df), ]
name.vec <- setNames(name.df$lineage_Condition_2, nm = name.df$ct_condition)
color.vec <- setNames(name.df$color, nm = name.df$ct_condition)

min_set_size = 1
font_size = 5
# Plot
plt <- ComplexUpset::upset(df, names(df), 
                           min_size = min_set_size,
                           set_sizes = FALSE, 
                           labeller = ggplot2::as_labeller(name.vec), 
                           name = NULL, 
                           height_ratio=0.8,
                           matrix=( intersection_matrix(geom=geom_point(size=0.75))), 
                    base_annotations = list(
                      # Barplot
                      'N of Genes with a TU-QTL' = intersection_size( fill = "#778da9",
                                                            text = list(size = 3/.pt, angle=90, vjust = 0.5, hjust = -0.2),
                                                            text_colors = c(on_background = "black",
                                                                            on_bar = "black")) +
                                      theme( panel.grid = element_blank(), axis.line = element_line(), 
                                             axis.title.y = element_text(size = font_size, vjust = -15),
                                             axis.text.y  = element_text(size = font_size))
                      ),
                    themes=upset_modify_themes(
                      list(
                        'intersection_size'=theme(text=element_text(size=font_size),
                                                  plot.margin = margin(0, 0, 0, 0, unit = "cm")),
                        'intersections_matrix'=theme(text=element_text(size=font_size),
                                                     plot.margin = margin(0, 0, 0, 0, unit = "cm"))
                        )), 
queries=list(upset_query(set= "T_CD4.NS", color = "#355070", fill="#355070"), # Not finding a systematic way of edditing colors !
             upset_query(set= "T_CD4.IAV", color = "#355070", fill="#355070"), 
             upset_query(set= "T_CD4.COV", color = "#355070", fill="#355070"), 
             upset_query(set= "T_CD8.NS", color = "#b56576", fill="#b56576"), 
             upset_query(set= "T_CD8.IAV", color = "#b56576", fill="#b56576"),
             upset_query(set= "T_CD8.COV", color = "#b56576", fill="#b56576"),
             upset_query(set= "MONO.NS", color = "#6d597a", fill="#6d597a"),
             upset_query(set= "MONO.IAV", color = "#6d597a", fill="#6d597a"),
             upset_query(set= "MONO.COV", color = "#6d597a", fill="#6d597a"),
             upset_query(set= "NK.NS", color = "#eaac8b", fill="#eaac8b"),
             upset_query(set= "NK.IAV", color = "#eaac8b", fill="#eaac8b"),
             upset_query(set= "NK.COV", color = "#eaac8b", fill="#eaac8b"), 
             upset_query(set= "B.NS", color = "#e56b6f", fill="#e56b6f"),
             upset_query(set= "B.IAV", color = "#e56b6f", fill="#e56b6f"), 
             upset_query(set= "B.COV", color = "#e56b6f",fill="#e56b6f")
             )
)

plt
file.name <- paste0("02_Plots/03_celltype-fractions/02_01_TU-Gene-ComplexUpset-01_lineage-min-set-", min_set_size, "-intersect-tested-genes.pdf") # for TU Gene
# file.name <- paste0("02_Plots/03_celltype-fractions/02_02_TU-QTL-ComplexUpset-01_lineage-min-set-", min_set_size, "-intersect-tested-genes.pdf") # for TU QTL
ggsave(file.name, plt, width = 18, height = 6, units = "cm")
# file.name <- gsub(".pdf", ".svg", file.name)
# ggsave(file.name, plt, width = 12, height = 5, units = "cm", bg = "white")
```


## Get numbers of lineage specific TU-QTLs

*For paper numbers*

```{r}
qtls.all <- read.table("01_Processed-Data/03_celltype-fractions/results_all_tested.tsv", sep = "\t", header = T)
colnames(qtls.all) <- plyr::revalue(colnames(qtls.all), c("cell.lineage" = "lineage", "condition" = "Condition"))
qtls.all$Condition <- plyr::revalue(qtls.all$Condition, c("Baseline" = "NS", "Influenza A Virus" = "IAV", "SARS-CoV-2" = "COV"))
# get list of snpIds tested in all lineage-Condtions
qtls.all <- qtls.all %>% filter(lineage %in% c("T.CD4", "T.CD8", "MONO")) %>% distinct(lineage, Condition, gene_name, snpId)

# Select those SNPs common to all lineage-Conditions
l <- split(qtls.all, qtls.all$lineage)
l <- lapply(l, function(x) split(x, x$Condition))
l <- unlist(l, recursive = F)
snp.l <- lapply(l, function(x) x$snpId)
common.snps <- Reduce(intersect, snp.l)
length(common.snps)

# N of lineage specific
qtls.sig <- read.table("01_Processed-Data/03_celltype-fractions/results_all_significant.tsv", sep = "\t", header = T)
qtls.sig <- qtls.sig %>% filter(lineage %in% c("T.CD4", "T.CD8", "MONO")) %>% distinct(lineage, Condition, gene_name, snpId)
qtls.sig <- qtls.sig[qtls.sig$snpId %in% common.snps, ]
qtls.sig <- qtls.sig %>% group_by(Condition, snpId) %>% summarise(n_lineage = n_distinct(lineage), 
                                                                  lineages = paste0(unique(lineage), collapse=","))

qtls.sig <- qtls.sig %>% group_by( Condition, n_lineage, lineages ) %>% summarise(n_qtls = n_distinct(snpId))
qtls.sig <- qtls.sig %>% group_by(Condition) %>% mutate(n_qtls_rel = n_qtls/ sum(n_qtls))
# qtls.sig <- qtls.sig %>% group_by(Condition, n_lineage) %>% summarise(n_qtls = sum(n_qtls), n_qtls_rel= sum(n_qtls_rel))
qtls.sig

# N of condition specific
qtls.sig <- read.table("01_Processed-Data/03_celltype-fractions/results_all_significant.tsv", sep = "\t", header = T)
qtls.sig <- qtls.sig %>% filter(lineage %in% c("T.CD4", "T.CD8", "MONO")) %>% distinct(lineage, Condition, gene_name, snpId)
qtls.sig <- qtls.sig[qtls.sig$snpId %in% common.snps, ]
qtls.sig <- qtls.sig %>% group_by(lineage, snpId) %>% summarise(n_conditions = n_distinct(Condition), 
                                                                conditions = paste0(unique(Condition), collapse=","))

qtls.sig <- qtls.sig %>% group_by( lineage, n_conditions, conditions ) %>% summarise(n_qtls = n_distinct(snpId))
qtls.sig <- qtls.sig %>% group_by(lineage) %>% mutate(n_qtls_rel = n_qtls/ sum(n_qtls))
# qtls.sig <- qtls.sig %>% group_by(lineage, n_conditions) %>% summarise(n_qtls = sum(n_qtls), n_qtls_rel= sum(n_qtls_rel))
qtls.sig
```

