---
title: "N-iso-gene"
author: "Ruben Chazarra Gil"
date: "`r Sys.Date()`"
output: html_document
---

# 0 Goal 

- Plotting Supplementary Fig. 2. A where we sxamine the nCount vs nFeature distribution for equivalence classes (ECs) and genes. 

# 0. Load functions
```{r}
source("../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../00_MetaData/03_Aesthetics-repo.R")
setwd("../04_EC-charac/07_nCount-nFeature/")
```

# 1. Calculate n_Count and n_Feature expressed 

*For each Equivalence class (EC) x cells object of each immune lineage (e.g. CD4+ T, CD8+ T, monocytes, NK, and B cells) which included cells from baseline, influenza A virus (IAV), and SARS-CoV-2 (COV)* 

*The same calculation was performed for the genes x cell object, displaying the same patter as for ECs, but only the nCount and nFeature EC values were finally plotted in Supplementary Fig. 2A*

```{r}
g <- read.RDS.files("../../02_Processing/03_Merge-split-cts/01_tcc/01_lineage/", pattern = "ECs-EUB-AFB.rds")
g <- lapply(g, function(seu){ # Recalculate nFeature and nCount
  meta = seu@meta.data
  mat = seu@assays$RNA@layers$counts
  n_count <- colSums(mat)
  mat@x[] <- 1 # replace all non zero values with 1 to calculate N feature
  n_feature <- colSums(mat)
  meta$nCount_RNA_post_fitler <- n_count
  meta$nFeature_RNA_post_fitler <- n_feature
  meta
})
g <- data.table::rbindlist(g)
saveRDS(g, "01_Processed-Data/01_Aquino-all-cells-metadata-01_ECs.rds")
# # Merge genes + ECs
# g <- readRDS("01_Processed-Data/01_Aquino-all-cells-metadata-01_genes.rds")
# e <- readRDS("01_Processed-Data/01_Aquino-all-cells-metadata-01_ECs.rds")
# meta <- plyr::ldply(list("gene" = g, "ecs" = e), .id = "class")
```


# 2. Plots

## 2.2 Boxplot By Condition

```{r}
# Merge genes + ECs
g <- readRDS("01_Processed-Data/01_Aquino-all-cells-metadata-01_genes.rds")
e <- readRDS("01_Processed-Data/01_Aquino-all-cells-metadata-01_ECs.rds")
meta <- plyr::ldply(list("gene" = g, "ecs" = e), .id = "class")

# process 
df <- meta %>% group_by(class, lineage, Condition) %>% summarise( mean_x = mean(nCount_RNA), mean_y = mean(nFeature_RNA), sd_x = sd(nCount_RNA), sd_y = sd(nFeature_RNA))

df$Condition <- factor(df$Condition, levels = names(condition.vec) )
df$Condition <- plyr::revalue(df$Condition, c("NS" = "Baseline")) # keep naming as: "Baseline"; "IAV"; "COV"
condition.color.vec <- c("Baseline" = "#969696", "IAV" = "#75A9D1", "COV" = "#c1121f")
color.vec <- setNames(lineage.df$color, nm = lineage.df$lineage)
df$lineage <- factor(df$lineage, levels = names(color.vec))
df$class <- factor(df$class, levels = c("gene", "ecs"))
# Only ECs
df <- df[df$class == "ecs", ]

plt <- ggplot(df, aes(x = mean_x, y = mean_y, color = Condition)) + 
  geom_point(size = 3/.pt) + 
  geom_errorbar(aes(ymin = mean_y - sd_y, ymax = mean_y + sd_y), size = 0.25) +
  geom_errorbarh(aes(xmin = mean_x - sd_x, xmax = mean_x + sd_x), size =  0.25) +
  scale_color_manual("", values = condition.color.vec ) +
  facet_grid( ~ lineage, scales = "free") + 
  xlab("UMI count") + ylab("EC count") + 
  labs(title = "Aquino All cells") + 
  plot_theme + theme(plot.title = element_blank(), #element_text(size = 8), 
                     axis.text.x = element_text(angle=30, hjust = 1, vjust = 1))
plt

file.name <- "02_Plots/02_nCount-nFeature-EC-genes-All-cells-boxplot-only-ECs.pdf"
ggsave(file.name, plot = plt, width = 10, height = 5, units = "cm")
```
