---
title: "04_popDTU-Validation"
author: "Ruben Chazarra Gil"
date: "`r Sys.Date()`"
output: html_document
---


# 0. Goal 

*Validation of Aquino results with different data* 

- Load functions
```{r}
source("../../../../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../../../../../01_Randolph2021/05_Differential-Analyses/02_DTU/00_Scripts/00_MANTA-functions.R")
source("../../../../../00_MetaData/03_Aesthetics-repo.R")
```

# 2. Overlap

`Rotival, M., Quach, H. & Quintana-Murci, L. Defining the genetic and evolutionary architecture of alternative splicing in response to infection. Nat Commun 10, 1671 (2019). https://doi.org/10.1038/s41467-019-09689-7`

### 2.1.1 Processing 

- Parse Rotival2019 population differential splicing data

NOTE: 
They do Differential Splicing with PSI between populations inside each stimulation group

- Load Supplementary Data 4
```{r}
require(readxl)
file.name <- "../../../../../../00_Bibliography/2019-Rotival-Nat_splicing-response-to-infection-Supp/2019-Rotival-Nat_splicing-response-to-infection-SupplementaryData_4.xlsx"
rotival <- read_excel(file.name, sheet = 1, skip = 18 )
str(rotival)

# manually determined by looking at excel sheet
common.cols <- names(rotival)[1:4]

# 1) Non-Stimullated
ns.cols <- names(rotival)[6:12]
ns <- rotival[ , c(common.cols, ns.cols) ]
# 2) LPS-Stimullated
lps.cols <- names(rotival)[14:20]
lps <- rotival[ , c(common.cols, lps.cols) ]
# 3) PAM
pam.cols <- names(rotival)[22:28]
pam <- rotival[ , c(common.cols, pam.cols) ]
# 4) R8
r8.cols <- names(rotival)[30:36]
r8 <- rotival[ , c(common.cols, r8.cols) ]
# 5) IAV-Stimullated
iav.cols <- names(rotival)[38:44]
iav <- rotival[, c(common.cols, iav.cols) ]
## collect to list
rotival.l <- list("NS" = ns, "LPS" = lps, "PAM" = pam, "R8" = r8, "IAV" = iav)

# update names
update.names <- c("AS_event_id", "gene_symbol", "event_type", "Is_DS_any_condition", 
                  "AS", "Delta_psi_AFB_EUB", "p_value_AFB_EUB", "FDR_AFB_EUB",
                  "DS_AFB_EUB", "sQTL_peak_SNP",
                  "frac_pop_differences_mediated_peak_SNP" )
rotival.l <- lapply(rotival.l, function(df) setNames(df, nm= update.names))

# Remove cases where differential splicing could not be determined confidently due to low expression (labelled in DS_AFB_EUB as "n.d")
rotival.l <- lapply(rotival.l, function(df) df[! df$DS_AFB_EUB %in% "n.d.", ])

# IMPORTANT: Rotival tests PSI, hence genes are repeated, keep one FDR value per Gene, the minimum
require(dplyr)
rotival.l <- lapply(rotival.l, function(df){
  as.data.frame( df %>% group_by(gene_symbol) %>% slice_min(FDR_AFB_EUB, with_ties = FALSE) ) 
})

# Rename certain columns to match ours
rotival.l <- lapply(rotival.l, function(x){
  colnames(x)[ match(c("gene_symbol", "FDR_AFB_EUB"), colnames(x)) ] <- c("gene_name", "fdr")
  return(x)
  })

# Save processed Rotival data
file.name <- "05_popDTU-Overlaps/01_Processed-Data/00_Rotival2019-population-DS-parsed.rds"
saveRDS(rotival.l, file.name)
```

### 2.1.2 Overlap

```{r}
dtu.l <- readRDS("01_Processed-Data/06_celltype-fractions/01_popDTU-unfiltered-list.rds")
# select 
# dtu.l <- select.df.from.nested.list(dtu.l, "POP")
# select lineage
dtu.l <- dtu.l$lineage
# restructure to Condition > celltype
dtu.l <- list("NS" = lapply(dtu.l, function(x) x[["NS"]]), 
              "IAV" = lapply(dtu.l, function(x) x[["IAV"]]), 
              "COV" = lapply(dtu.l, function(x) x[["COV"]])
              )
# Add gene names
# tx2gene = "../../../../../01_Quantif/00_IndexData/02_GENCODEv42_filt-IAV-SARS_Cov_2-tx2gene-with-ENS.tsv"
# tx2gene <- read.table( tx2gene, header = T, sep = "\t", na.strings = "" )
# dtu.l <- add.gene_name.2.gene_id(dtu.l,  tx2gene, gene_id.col = "gene_id",  gene_name.col = "gene_name")
```

- Overlap
```{r}
overlap.l <- lapply(dtu.l, function(condition_1){
  lapply(condition_1, function(celltype){
    lapply(rotival.l, function(condition_2){
        test.overlap.pro(m1 = celltype, fdr.col = "fdr", fdr.thres.1 = 0.05, gene.col.1 = "gene_name",
                       FC.col.1 = NA, FC.thres.1 = NA, 
                       m2 = condition_2, fdr.col.2 = "fdr", fdr.thres.2 = 0.05, gene.col.2 = "gene_name", 
                       FC.col.2 = "Delta_psi_AFB_EUB", FC.thres.2 = 0.05) # 
    })
  })
})

# Merge
overlap.l <- lapply(overlap.l, function(condition_1) lapply(condition_1, function(celltype) plyr::ldply(celltype, .id = "Rotival")))
overlap.l <- lapply(overlap.l, function(condition_1) plyr::ldply(condition_1, .id = "celltype"))  
df = plyr::ldply(overlap.l, .id = "Aquino")
head(df)
```

- Save
```{r}
file.name <- "05_popDTU-Overlaps/01_Processed-Data/06_celltype-fractions/00_popDTU-Aquino2023-vs-Rotival2019.rds"
saveRDS(df, file.name)
```

- Add significance label
```{r}
require(dplyr)
# Process df to include y coordinate of label 
df$model.2.specific <- NULL # model_2 genes will appear in another plot
df.i <- reshape2::melt(df[ , c("Aquino", "Rotival", "celltype", "shared", "model.1.specific" )]) %>% group_by(celltype, Aquino,Rotival) %>% mutate(label_y = sum(value)) %>% ungroup()
# add p_value
df = merge(df.i,
      df [, c("celltype", "Aquino", "Rotival", "p_value") ], 
      by = c("celltype", "Aquino", "Rotival"))

df = add.p_value.label(df)
df
```

### 2.1.3 Plots

##### OR

```{r}
# Read data
file.name <- "05_popDTU-Overlaps/01_Processed-Data/06_celltype-fractions/00_popDTU-Aquino2023-vs-Rotival2019.rds"
df <- readRDS( file.name )
# add p_value_label
df = add.p_value.label(df)
# subset to MONOCYTES
df <- df[ df$celltype == "MONO", ]
# subset to Monocytes and  comparisons of interest 
df <- df [  df$celltype == "MONO" & df$Aquino  == "NS" & df$Rotival  == "NS" | df$Aquino  == "IAV" & df$Rotival  == "IAV" | df$Aquino  == "COV" & df$Rotival  == "IAV",   ]

# Plot
df$Condition <- plyr::revalue(df$Aquino, condition.vec.short)
df$Condition <- factor(df$Condition, levels = rev(condition.vec.short))                         
color.vec <- condition.color.vec.short

odds.plot <- ggplot(data = df, aes(x = odds.ratio, y = Condition, color = Condition)) + 
  geom_point(size = 1) + 
  geom_text(aes(x = odds.ratio, y = Condition, label = p_value_label), color = "black", vjust = -0.5, size = 3.5/.pt) + 
  geom_vline(aes(xintercept = 1), size = .5, linetype = "dashed", linewidth =0.25) +
  geom_errorbarh(aes(xmax = conf.int.upper, xmin = conf.int.lower), size = .5,height = .2, linewidth =0.25) + 
  scale_color_manual(guide = "none", values = color.vec) + 
  labs(title = "tit", subtitle = "sub_tit") + ylab("y_lab") + xlab("Odds Ratio") + 
  plot_theme + theme(axis.title.y = element_blank())
  # theme(panel.background = element_blank(), plot.background = element_blank(), 
  #       axis.line.x = element_line(), axis.line.y = element_line(),
  #       axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5), 
  #       axis.text.y = element_text(size = 10, angle = 0, hjust = 0.5), axis.title.y = element_blank(), 
  #       legend.position = "none")
odds.plot
# Save
file.name <- "05_popDTU-Overlaps/02_Plots/06_celltype-fractions/02_overlap-Rotival-only-MONO-01_OR.pdf" 
ggsave(file.name, odds.plot, width = 4, height = 2.5, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, odds.plot, width = 4, height = 2.5, units = "cm", bg = "white")
```

##### BarPlot
```{r}
# Read data
file.name <- "05_popDTU-Overlaps/01_Processed-Data/06_celltype-fractions/00_popDTU-Aquino2023-vs-Rotival2019.rds"
df <- readRDS( file.name )
# add p_value_label
df = add.p_value.label(df)
# subset to MONOCYTES
df <- df[ df$celltype == "MONO", ]
# subset to Monocytes and  comparisons of interest 
df <- df [  df$celltype == "MONO" & df$Aquino  == "NS" & df$Rotival  == "NS" | df$Aquino  == "IAV" & df$Rotival  == "IAV" | df$Aquino  == "COV" & df$Rotival  == "IAV",   ]

# Factor
df$Condition <- factor(df$Aquino, levels = rev(names(condition.vec.short)))
df$Condition <- plyr::revalue(df$Condition, condition.vec.short) 
color.vec <- condition.vec.short

df <- reshape2::melt( df[, c("Condition", "shared", "model.1.specific", "model.2.specific") ], 
                      value.name = "n_genes", variable.name = "model")
df$model <- plyr::revalue(df$mode, c("model.1.specific" = "Aquino 2023", "model.2.specific" = "Rotival 2019")) 
color.vec <- c("Aquino 2023" = "#4a4e69", "shared" = "#f2e9e4", "Rotival 2019" = "#c9ada7")
df$model <- factor(df$model, levels = c("Aquino 2023", "shared", "Rotival 2019"))

plt <- ggplot(data = df, aes(x = n_genes, y = Condition, fill = model)) + 
  geom_bar(stat="identity", color = "black", linewidth=0.25, position = position_stack())  +
  geom_text(aes(x = n_genes, label = n_genes), vjust=0.5, color= "black", size=3.5/.pt,  
            position = position_stack(vjust = .5 )) +
  # geom_text(aes(y = Condition, x = label_x, label = p_value_label),angle = 90, 
  #           vjust = 1.25, hjust = 0.5, size = 3) +  # p_value
  scale_fill_manual("", values = color.vec,  drop = F) +
  labs(title = "Population DTU sharing", subtitle = "Aquino Population DTU vs ROtival 2019") +
  xlab("N Population DTU Genes") + 
  plot_theme + 
  theme(axis.title.y = element_blank())
  # theme(panel.background = element_blank(), plot.background = element_blank(), 
  #       axis.line.x = element_line(), axis.line.y = element_line(),
  #       axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5), 
  #       axis.text.y = element_text(size = 10, angle = 0, hjust = 0.5), axis.title.y = element_blank()) + 
  # guides(fill = guide_legend(reverse=TRUE))

# Save
file.name <- "05_popDTU-Overlaps/02_Plots/06_celltype-fractions/02_overlap-Rotival-only-MONO-02_BarPlot.pdf" 
ggsave(file.name, plt, width = 5, height = 2.5, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, width = 5, height = 2.5, units = "cm", bg = "white")
```


<!-- ### 2.3.4 Rotival popDS DS sharing across conditions -->

<!-- *2025-12-19* We see that popDTU genes are highlighy condition-specific (e.g Baseline, IAV, COV), I want to see Rotival 2019 popDS sharing across conditions* -->

<!-- ```{r} -->
<!-- df.l = readRDS("05_popDTU-Overlaps/01_Processed-Data/00_Rotival2019-population-DS-parsed.rds") -->
<!-- df = plyr::ldply(df.l, .id = "Condition") -->
<!-- # Filter popDS -->
<!-- "Delta_psi_AFB_EUB" -->

<!-- # Filter by PSI effect size and FDR -->
<!-- df <- df[ abs(df$Delta_psi_AFB_EUB) > 0.05 & df$fdr < 0.05, ] -->
<!-- # Condition sharing -->
<!-- df <- df %>% group_by(gene_name) %>% summarise(n_conditions = n_distinct(Condition)) -->
<!-- df <- df %>% group_by(n_conditions) %>% summarise(n_genes = n_distinct(gene_name)) -->
<!-- df <- df %>% mutate(n_genes_rel = n_genes / sum(n_genes)) -->
<!-- ``` -->



