---
title: "03_DTU-Viz"
author: "Ruben Chazarra Gil"
date: "`r Sys.Date()`"
output: html_document
---

## 0. Goal 

*Supplementary Fig. 14. Validation of population-level differential transcript usage (DTU) results by permutation and comparison with previous studies. *
- a) Permutation analysis of population DTU. For each immune lineage (e.g., CD4⁺ T cells) and condition (e.g., baseline), donor labels were randomly shuffled and the DTU analysis was repeated 100 times. Boxplots show the number of significant population DTU genes detected under permutation (FDR < 0.05, Benjamini–Hochberg correction); grey points indicate outliers. 
 
- Load data 
```{r}
source("../../../../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../../../../00_MetaData/03_Aesthetics-repo.R")
setwd("../03_Differential/02_DTU/02_DTU/02_popDTU/02_DA/")
```

# Supp. Fig. 14.A

### Process
- Permuted data
```{r}
file.l <- list("lineage" = read.RDS.files(path = "../01_popDTU/06_celltype_fractions/01_lineage/", pattern = "permutations-x100.rds")))

count_sig_permuted <- function(df){
  ## Filter a dataframe by fdr.col{
  if(class(df) == "list"){ lapply(df, count_sig_permuted )
    } else {df %>% summarise(n_genes = sum(fdr < 0.05)) }
}

dtu.l <- count_sig_permuted(file.l)
# Process
# Merge By covariate
dtu.l = lapply(dtu.l, function(l) lapply(l, function(celltype) lapply(celltype, function(it) lapply(it, function(condition) plyr::ldply(condition, .id = "covariate")))))
## By Condition
dtu.l = lapply(dtu.l, function(l) lapply(l, function(celltype) lapply(celltype, function(it) plyr::ldply(it, .id = "Condition"))))
# By iteration
dtu.l = lapply(dtu.l, function(l) lapply(l, function(celltype) plyr::ldply(celltype, .id = "iteration")))
# By celltype
dtu.l = lapply(dtu.l, function(l) plyr::ldply(l, .id = "celltype"))
# Save count
saveRDS(dtu.l, "01_Processed-Data/06_celltype-fractions/01_popDTU-permutation-n-significant-DTU-count.rds")
```

### Boxplot only permutations
```{r}
dtu.l <- readRDS("01_Processed-Data/06_celltype-fractions/01_popDTU-permutation-n-significant-DTU-count.rds")
dtu <- dtu.l$lineage
df <- dtu[dtu$covariate %in% "POP", ]
df$Condition <- factor(df$Condition, levels = names(condition.vec.short))
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)
df$celltype <- factor(df$celltype, levels = rev(lineage.df$lineage))

plt <- ggplot(data = df, aes(y = celltype)) + 
  geom_boxplot(aes(x = n_genes), fill = "grey", outlier.colour = "grey50", outlier.size = 0.5, linewidth = 0.25 )+ 
  facet_grid( Condition ~ ., scales = "free_x", switch = "y") +
  scale_color_manual("Lineage", values = lineage.color.vec, drop = F) +
  labs(title = "tit", subtitle = "sub_tit") + xlab("Permutation (N=100) population DTU genes") + ylab("y_lab") + 
  plot_theme + theme( axis.title.y = element_blank(), axis.title.x = element_text(hjust = 1)) + 
  guides(fill = FALSE, color = FALSE ) 
# Save
file.name <- "02_Plots/06_celltype-fractions/06_popDTU-permutations-x100-01_lineage-POPOnly-by-Condition.pdf"
ggsave(file.name, plt, width = 4.25, height = 5, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, width = 4.25, height = 5, units = "cm", bg = "white")
```



