---
title: "infectionDE-Enrichment"
author: "Ruben Chazarra Gil"
date: "`r Sys.Date()`"
output: html_document
---

# 0. Goal 

*Supplementary Fig. 8*. Overrepresentation analysis (ORA) of infection-associated DTU genes in response to SARS-CoV-2 (COV) across immune lineages. ORA was performed for different DTU gene sets: all DTU genes (DTU); DTU genes overlapping with differentially expressed genes (DTU–DGE); and DTU–DGE genes further stratified into upregulated (DTU–DGE Up) and downregulated (DTU–DGE Down) categories. Shown are Gene Ontology (GO) biological process terms with FDR < 0.05. Point color represents –log₁₀(FDR), and point size reflects the enrichment odds ratio.

AND

*Supplementary Fig. 9*. Overrepresentation analysis (ORA) of infection-associated DTU genes in response to influenza A (IAV) across immune lineages. ORA was performed for different DTU gene sets: all DTU genes (DTU); DTU genes overlapping with differentially expressed genes (DTU–DGE); and DTU–DGE genes further stratified into upregulated (DTU–DGE Up) and downregulated (DTU–DGE Down) categories. Shown are Gene Ontology (GO) biological process terms with FDR < 0.05. Point color represents –log₁₀(FDR), and point size reflects the enrichment odds ratio.


- Load packages
```{r}
source("../../../../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../../../../00_MetaData/03_Aesthetics-repo.R")
setwd("../03_Differential/02_DTU/02_DTU/01_infectionDTU/02_DA/")

suppressPackageStartupMessages(require(enrichplot)) 
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(scales))
suppressPackageStartupMessages( require(rrvgo) )
suppressPackageStartupMessages(require(org.Hs.eg.db))
```



# 1. Data Processing

## 1.1 Restructure
- Get significant ORA results, and restructure data for plotting

```{r}
dtu.ora <- readRDS("04_infectionDTU-ORA/01_Processed-Data/05_Filt-1-5-0.8-0.01/01_infectionDTU-ORA-clusterProfiler-GO-01_lineage.rds")

get.result.ora <- function ( df, fdr.thres ) {
  if ( class(df) == "list" ) { lapply (df, function(df) get.result.ora(df, fdr.thres))
  } else { # class == "enrichResult"
    if(!is.null(df)){
    df = df@result # get data.frame
    df = df[ df$p.adjust < fdr.thres, ]
    return(df)
    }
  }
}

dtu.ora = get.result.ora(dtu.ora,fdr.thres = 0.05)

# Restructure to cell type > database > condition > subset
# 1. Restucture to cell type > condition > database > subset
dtu.ora = lapply(dtu.ora, function(ct)
  lapply(ct, function(cond){
    list(
      "BP" = lapply(cond, function(x) x[["BP"]]) # , 
      # "MF" = lapply(cond, function(x) x[["MF"]]), # excluding these databases in last run
      # "CC" = lapply(cond, function(x) x[["CC"]])
      )
  }))

# 2. Restucture to cell type > database > condition > subset
dtu.ora = lapply(dtu.ora, function(ct){
    list(
      "BP" = lapply(ct, function(x) x[["BP"]])#, 
      #"MF" = lapply(ct, function(x) x[["MF"]]),
      #"CC" = lapply(ct, function(x) x[["CC"]])
      )
  })

# Merge
## By set
dtu.ora <- lapply(dtu.ora, function(ct)
  lapply(ct, function(db)
    lapply(db, function(cond) plyr::ldply(cond, .id = "set"))
  ))

## By Condition
dtu.ora <- lapply(dtu.ora, function(ct)
  lapply(ct, function(db) plyr::ldply(db, .id = "Condition"))
  )

## unlist 1 level 
dtu.ora <- unlist(dtu.ora, recursive = F)
# remove  empty
dtu.ora <- dtu.ora[ unlist(lapply(dtu.ora, nrow)) > 0 ]

# Save grouped data
file.name <- "04_infectionDTU-ORA/01_Processed-Data/05_Filt-1-5-0.8-0.01/02_infectionDTU-ORA-filtered-grouped-01_lineage.rds"
saveRDS( dtu.ora, file.name )
```

## 1.2 Add Odds Ratio to ORA results based on GeneRatio and BgRatio values 

```{r}
# Add Odds Ratio (and 95% CI) to clusterProfiler enrichGO results
add_enrichment_or <- function(enrich_res, method = c("fisher", "closed_form")) {
  if(class(enrich_res) == "list"){ 
    lapply(enrich_res, function(x) add_enrichment_or(x, method = c("fisher", "closed_form")))
  }else{
  method <- match.arg(method)
  df <- as.data.frame(enrich_res)

  if (!all(c("GeneRatio", "BgRatio", "Count") %in% names(df))) {
    stop("Expected columns 'GeneRatio', 'BgRatio', and 'Count' not found.")
  }

  # Parse ratios safely
  parse_ratio_num <- function(x, idx) {
    # idx=1 numerator, idx=2 denominator
    as.numeric(vapply(strsplit(x, "/", fixed = TRUE), function(z) z[[idx]], FUN.VALUE = character(1L)))
  }

  k <- as.numeric(df$Count)                                   # overlap in query ∩ term
  n <- parse_ratio_num(df$GeneRatio, 2L)                      # size of query
  M <- parse_ratio_num(df$BgRatio, 1L)                        # term size in background
  N <- parse_ratio_num(df$BgRatio, 2L)                        # background size

  # Sanity checks to catch inconsistencies early
  bad <- is.na(k) | is.na(n) | is.na(M) | is.na(N) | k > n | k > M | n > N | M > N
  if (any(bad)) {
    warning("Some rows have inconsistent counts; OR set to NA for those rows.")
  }

  # Compute ORs
  or <- rep(NA_real_, nrow(df))
  ci_lo <- rep(NA_real_, nrow(df))
  ci_hi <- rep(NA_real_, nrow(df))

  for (i in seq_len(nrow(df))) {
    if (bad[i]) next

    a <- k[i]
    b <- n[i] - k[i]
    c <- M[i] - k[i]
    d <- N[i] - M[i] - b

    if (any(c(a, b, c, d) < 0)) {
      next
    }

    if (method == "fisher") {
      # Why: fisher.test gives finite OR + exact CI, handles zeros better than closed-form
      mat <- matrix(c(a, b, c, d), nrow = 2,
                    dimnames = list(c("Query", "NotQuery"), c("InTerm", "NotInTerm")))
      ft <- suppressWarnings(fisher.test(mat, alternative = "greater"))
      or[i] <- unname(ft$estimate)
      ci_lo[i] <- ft$conf.int[1]
      ci_hi[i] <- ft$conf.int[2]
    } else {
      # Closed-form; may be Inf with zeros
      num <- a * d
      den <- b * c
      if (den == 0) {
        or[i] <- Inf
        ci_lo[i] <- NA_real_
        ci_hi[i] <- NA_real_
      } else {
        or[i] <- num / den
        ci_lo[i] <- NA_real_
        ci_hi[i] <- NA_real_
      }
    }
  }

  df$odds_ratio <- or
  df$or_ci_low  <- ci_lo
  df$or_ci_high <- ci_hi
  df
  }
}
# Run OR
file.name <- "04_infectionDTU-ORA/01_Processed-Data/05_Filt-1-5-0.8-0.01/02_infectionDTU-ORA-filtered-grouped-01_lineage.rds"
dtu.ora <- readRDS(file.name)
dtu.ora <-  add_enrichment_or(dtu.ora, method="fisher")
saveRDS(dtu.ora, file.name)
```


# Supp. Fig. 8. /9. ORA for 

*Supplementary Fig. 8. Overrepresentation analysis (ORA) of infection-associated DTU genes in response to SARS-CoV-2 (COV) across immune lineages. ORA was performed for different DTU gene sets: all DTU genes (DTU); DTU genes overlapping with differentially expressed genes (DTU–DGE); and DTU–DGE genes further stratified into upregulated (DTU–DGE Up) and downregulated (DTU–DGE Down) categories. Shown are Gene Ontology (GO) biological process terms with FDR < 0.05. Point color represents –log₁₀(FDR), and point size reflects the enrichment odds ratio.*

- Load data
```{r}
file.name <- "04_infectionDTU-ORA/01_Processed-Data/05_Filt-1-5-0.8-0.01/02_infectionDTU-ORA-filtered-grouped-01_lineage.rds"
dtu.ora <- readRDS( file.name )
# Rename
names(dtu.ora) <-  gsub(".BP", "", names(dtu.ora))
## Merge By lineage
dtu.ora <- plyr::ldply(dtu.ora, .id = "lineage")
dtu.ora <- dtu.ora[dtu.ora$Condition %in% c("IAV.NS", "COV.NS"), ]
  
df <- dtu.ora
df$lineage <- factor(df$lineage, levels = lineage.df$lineage)
df$Condition <- factor(df$Condition, levels = names(condition.pairwise.vec.short))
df$Condition <- plyr::revalue(df$Condition, condition.pairwise.vec.short)
df$set <- factor(df$set, levels = c("DTU.all", "DTU.unique", "DTU.DGE", "DTU.DGE.Up", "DTU.DGE.Down"))
df$set <- plyr::revalue(df$set, c("DTU.all" = "DTU", "DTU.unique" = "DTU unique",
                                    "DTU.DGE" = "DTU-DGE", "DTU.DGE.Up" = "DTU-DGE Up", "DTU.DGE.Down" ="DTU-DGE Down"))

# subset
# df <- df[df$p.adjust < 0.05 & df$Condition %in% "IAV-Baseline", ]
df <- df[df$p.adjust < 0.05 & df$Condition %in% "COV-Baseline", ]

# Plot
plt <- ggplot(df, aes(x = set, y = Description, fill = -log10(p.adjust))) + # color by log10FDR
  geom_point( shape=21, aes(size = odds_ratio)) +
  facet_grid( Condition ~ lineage, scales = "free",  space = "free", switch = "y") +
  labs(title = "ORA BP", subtitle = "FDR < 0.05") + #xlab("Cell type") +
  scale_fill_viridis_c(name = "-log10(FDR)") +
  scale_y_discrete(labels = function(x) stringr::str_trunc(x, 40), limits=rev ) +
  scale_size_continuous("Odds Ratio", range = c(1, 3.5), breaks = c(2, 5, 10, 15, 20, 60)) +
  scale_x_discrete(drop=T) +
  plot_theme + 
      theme(axis.title.y = element_blank(),
          axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
          axis.title.x = element_blank(),
          axis.text.y = element_text(size = 5),
          panel.background = element_rect(fill = "white"),
          axis.line.x = element_line(), axis.line.y = element_line(),
          strip.background = element_rect(fill = NA, color = "black", size = 1),
          panel.grid = element_line(color = "grey", linewidth = 0.05),
          strip.placement = "outside")

# Save
## Supplementary Fig 8. 
file.name <- "04_infectionDTU-ORA/02_Plots/05_Filt-1-5-0.8-0.01/02_02_infection-ORA-all-celltypes-01_lineage-COV.NS_size-by-OR.pdf"
## Supplementary Fig 9.
# file.name <- "04_infectionDTU-ORA/02_Plots/05_Filt-1-5-0.8-0.01/02_02_infection-ORA-all-celltypes-01_lineage-IAV.NS_size-by-OR.pdf"
ggsave(file.name, plt, height = 20, width = 20, units="cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, height = 20, width = 20,  units = "cm", bg = "white")
```
