---
title: "01_DA"
author: "Ruben Chazarra Gil"
date: "`r Sys.Date()`"
output: html_document
---

# 0. Goal

*Supplementary Fig. 17.*  Population-associated differential transcript usage (DTU) genes with the largest ancestry differences are more likely to be under genetic regulation and associated with large-effect transcript usage quantitative trait loci (TU-QTL). 
- a) Proportion of population DTU genes harboring a TU-QTL, stratified by TU effect size quartiles (maximum absolute difference in mean EC ratio across sample groups, MD value), showing enrichment of TU-QTLs among genes with larger ancestry-associated TU differences. 
- b) Correlation between TU-QTL effect size (MD value) and DTU effect size (MD value) for population DTU genes with at least one TU-QTL, across immune lineages and conditions. Spearman correlation coefficients (R) and corresponding p values are shown. 


- Load functions
```{r}
source("../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../../01_Randolph2021/05_Differential-Analyses/02_DTU/00_Scripts/00_MANTA-functions.R")
source("../../00_MetaData/03_Aesthetics-repo.R")
setwd("../06_sQTL/03_DA/")
```


# Supp. Fig 17.A 

*Supplementary Fig. 17.*  a) Proportion of population DTU genes harboring a TU-QTL, stratified by TU effect size quartiles (maximum absolute difference in mean EC ratio across sample groups, MD value), showing enrichment of TU-QTLs among genes with larger ancestry-associated TU differences. 

```{r}
# TU QTLs
qtl <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
select.qtl <- c("lineage", "Condition", "geneId", "gene_name", "snpId", "best.snp", "md", "fdr")
qtl <- qtl[, select.qtl]
# Keep best, we only want to know if a popDTU has at least one TU-QTL
qtl <- qtl[qtl$snpId == qtl$best.snp, ]

# Pop DTU
dtu <- readRDS("../../03_Differential/02_DTU/02_DTU/02_popDTU/02_DA/01_Processed-Data/06_celltype-fractions/01_popDTU-filtered-df.rds")
dtu <- dtu$lineage
select.dtu <- c("celltype", "Condition", "gene_id", "gene_name", "MD", "fdr")
dtu <- dtu[, select.dtu ]
dtu$MD <- as.numeric(dtu$MD)
# Add MD Quantiles by lineage-Condition 
dtu$celltype_Condition <- paste0(dtu$celltype, "_", dtu$Condition)
dtu.l <- split(dtu, dtu$celltype_Condition)
dtu.l <- lapply(dtu.l, function(x){
  q.vec =  quantile(x$MD, probs = seq(0, 1, 0.25), include_lowest  = T)
  x$MD_quantile <-  cut(x$MD, breaks = q.vec, labels = paste0("Q", c(1:4)),  include.lowest= T)
  x
})
dtu <- data.table::rbindlist(dtu.l)

# Merge
df <- merge(dtu, qtl,
            by.x = c("celltype", "Condition", "gene_id", "gene_name"), 
            by.y = c("lineage", "Condition", "geneId", "gene_name"), 
            all.x = T, all.y = F)
# Label genes with
df$is.QTL <- ifelse(is.na(df$snpId), FALSE, TRUE)

# Stratify popDTU by MD
df <- df %>% group_by(celltype, Condition, MD_quantile) %>% summarise(n_genes_qtl = sum(is.QTL == "TRUE"),
                                                                      n_genes = n_distinct(gene_id))
df$n_genes_qtl_rel  = df$n_genes_qtl / df$n_genes

# Plot
df$bar_label <- paste0(df$n_genes_qtl, "/", df$n_genes)
df$Condition <- factor(df$Condition, levels = names(condition.vec.short) )
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)
color.vec <- setNames(lineage.df$color, nm = lineage.df$lineage)
df$celltype <- factor(df$celltype, levels = rev(names(color.vec)))
df$MD_quantile <- factor(df$MD_quantile, levels = rev(c("Q1", "Q2", "Q3", "Q4")))
alpha.vec <- c("Q1" = "Very Low", "Q2" = "Low", "Q3" = "Mid", "Q4" = "High")

# Plotce
plt <- ggplot(data = df, aes(y = celltype, x = n_genes_qtl_rel, fill = celltype)) +
  geom_bar( data = df, aes(alpha = MD_quantile), stat = "identity", linewidth=0.25, width = 0.8, position_dodge( width=0.8 )) +
  geom_text(aes(x = n_genes_qtl_rel, label = bar_label, group = MD_quantile ),
            alpha = 1, color = "black", size = 3/.pt, hjust = -0.1, vjust = 0.5, position = position_dodge( width=0.8 )) +
  facet_grid(Condition ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_fill_manual( values = lineage.color.vec ) +
  scale_alpha_manual("Effect size", values = rev(c(0.4, 0.6, 0.8, 1)), labels = alpha.vec) +
  xlab("PopDTU genes with a TU-QTL (%)") + ylab("y_lab") + 
  plot_theme + theme(axis.title.y = element_blank(), 
                     legend.position=c(0.86, 0.5)) + 
  guides(fill = FALSE, color = FALSE, alpha = guide_legend(reverse = T))
plt

# Save
file.name <- paste0("02_Plots/03_celltype-fractions/03_03_PopDTU-fraction-with-QTL-stratified-by-MD_01.pdf")
ggsave(file.name, plt, width = 5.5, height = 7, units = "cm")
file.name <- gsub(".pdf", ".svg", file.name)
ggsave(file.name, plt, width = 5.5, height = 7, units = "cm", bg = "white")
```


# Supp. Fig 17.B 

*Supplementary Fig. 17.* b) Correlation between TU-QTL effect size (MD value) and DTU effect size (MD value) for population DTU genes with at least one TU-QTL, across immune lineages and conditions. Spearman correlation coefficients (R) and corresponding p values are shown. 

```{r}
# TU QTLs
qtl <- readRDS("01_Processed-Data/03_celltype-fractions/01_sQTLseekeR2-permuted-fdr-0.01-df.rds")
# maybe we could select best SNP
select.qtl <- c("lineage", "Condition", "geneId", "gene_name", "snpId", "best.snp", "md", "fdr")
qtl <- qtl[, select.qtl]

# Pop DTU
dtu <- readRDS("../../03_Differential/02_DTU/02_DTU/02_popDTU/02_DA/01_Processed-Data/06_celltype-fractions/01_popDTU-filtered-df.rds")
dtu <- dtu$lineage
select.dtu <- c("celltype", "Condition", "gene_id", "gene_name", "MD", "fdr")
dtu <- dtu[, select.dtu ]
dtu$MD <- as.numeric(dtu$MD)

df <- merge(dtu, qtl,
             by.x = c("celltype", "Condition", "gene_id", "gene_name"), 
             by.y = c("lineage", "Condition", "geneId", "gene_name"), 
             all.x = T, all.y = F)

# some sQTLseekeR2 md values are NA, remove
df <- df[!is.na(df$md), ]

# Select best SNP 
# df <- df[df$snpId == df$best.snp,  ]

# Plot 
df$Condition <- factor(df$Condition, levels = names(condition.vec.short) )
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)
color.vec <- setNames(lineage.df$color, nm = lineage.df$lineage)
df$celltype <- factor(df$celltype, levels = lineage.df$lineage)

plt <- ggplot(data = df, aes(x = md, y = MD, color = celltype )) +
  # xlim(0, 0.5) + ylim(0, 0.5) + 
  geom_point(shape = 20, size = 0.05, alpha = 1)  + 
  geom_smooth(aes(fill = celltype), method = "lm", size = 0.5) +  # Add trend line
  ggpubr::stat_cor(aes(color = celltype), method = "spearman", # for all the points
                   label.x = 0, label.y = 0.2, size = 1.5) +
  facet_grid(  Condition ~ celltype, switch = "y") +
  scale_color_manual( "", values = lineage.color.vec ) +
  scale_fill_manual( "", values = lineage.color.vec ) +
  # scale_x_continuous(breaks = c(0, 0.2, 0.4)) + 
  # scale_y_continuous(breaks = c(0, 0.2, 0.4)) + 
  labs(title = "tit", subtitle = "sub_tit") + xlab("Population DTU gene MD value") + ylab("TU-QTL MD value") + 
  plot_theme + 
  guides( fill = FALSE, color = FALSE)
plt
file.name <- paste0("02_Plots/03_celltype-fractions/03_02_MD-corr-TU-QTL-PopDTU_01.pdf")
ggsave(file.name, plt, width = 11, height = 7, units = "cm")
file.name <- gsub(".pdf", ".svg",  file.name)
ggsave(file.name, plt, width = 11, height = 7, units = "cm", bg = "white")
```
