---
title: "popDE-Enrichment"
author: "Ruben Chazarra Gil"
date: "`r Sys.Date()`"
output: html_document
---

## 0. Goal 


Supplementary Fig. 13. Pathway enrichment analysis of population-associated differential transcript usage (DTU) genes.Overrepresentation analysis (ORA) of population DTU genes identified across immune lineages under baseline, influenza A virus (IAV), and SARS-CoV-2 (COV) conditions. Gene sets include: (i) all DTU genes (DTU); (ii) DTU genes overlapping with differentially expressed genes (DTU & DGE); and (iii) DTU & DGE genes further stratified as upregulated in Africans (DTU & DGE AFR) or upregulated in Europeans (DTU & DGE EUR). Gene Ontology (GO) biological process terms with FDR < 0.05 (Benjamini–Hochberg correction) are shown. GO terms were grouped using rrvgo4 under a similarity threshold of 0.7. Dot color intensity and size represent the fraction and absolute number of GO terms within each grouped category, respectively.


 - Load functions
```{r}
source("../../../../../../01_Randolph2021/05_Differential-Analyses/00_Scripts/00_DA-Functions.R")
source("../../../../../00_MetaData/03_Aesthetics-repo.R")
setwd("../03_Differential/02_DTU/02_DTU/02_popDTU/02_DA/")

suppressPackageStartupMessages(require(enrichplot)) # cannot install this in local!
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(scales))
# Run rrvgo
suppressPackageStartupMessages( require(rrvgo) )
suppressPackageStartupMessages(require(org.Hs.eg.db))
```


# 1. Data Processing

## 1.1 Restructure

- Load data
```{r}
dtu.ora <- readRDS("04_popDTU-ORA/01_Processed-Data/06_celltype-fractions/01_popDTU-ORA-clusterProfiler-GO-BP-01_lineage.rds")

get.result.ora <- function ( df, fdr.thres ) {
  if ( class(df) == "list" ) { lapply (df, function(df) get.result.ora(df, fdr.thres))
  } else { # class == "enrichResult"
    if(!is.null(df)){
    df = df@result # get data.frame
    df = df[ df$p.adjust < fdr.thres, ]
    return(df)
    }
  }
}

dtu.ora = get.result.ora(dtu.ora,fdr.thres = 0.05)

# Restructure to cell type > database > condition > subset
# 1. Restucture to cell type > condition > database > subset
dtu.ora = lapply(dtu.ora, function(ct)
  lapply(ct, function(cond){
    list(
      "BP" = lapply(cond, function(x) x[["BP"]]) # , 
      # "MF" = lapply(cond, function(x) x[["MF"]]), # excluding these databases in last run
      # "CC" = lapply(cond, function(x) x[["CC"]])
      )
  }))

# 2. Restucture to cell type > database > condition > subset
dtu.ora = lapply(dtu.ora, function(ct){
    list(
      "BP" = lapply(ct, function(x) x[["BP"]])#, 
      #"MF" = lapply(ct, function(x) x[["MF"]]),
      #"CC" = lapply(ct, function(x) x[["CC"]])
      )
  })

# Merge
## By set
dtu.ora <- lapply(dtu.ora, function(ct)
  lapply(ct, function(db)
    lapply(db, function(cond) plyr::ldply(cond, .id = "set"))
  ))

## By Condition
dtu.ora <- lapply(dtu.ora, function(ct)
  lapply(ct, function(db) plyr::ldply(db, .id = "Condition"))
  )

## unlist 1 level 
dtu.ora <- unlist(dtu.ora, recursive = F)
# remove  empty
dtu.ora <- dtu.ora[ unlist(lapply(dtu.ora, nrow)) > 0 ]

# Save grouped data
file.name <- "04_popDTU-ORA/01_Processed-Data/06_celltype-fractions/02_popDTU-ORA-filtered-grouped-01_lineage.rds"
saveRDS( dtu.ora, file.name )
```

## 1.2 Add Odds Ratio to ORA results based on GeneRatio and BgRatio values 

```{r}
# Add Odds Ratio (and 95% CI) to clusterProfiler enrichGO results
add_enrichment_or <- function(enrich_res, method = c("fisher", "closed_form")) {
  if(class(enrich_res) == "list"){ 
    lapply(enrich_res, function(x) add_enrichment_or(x, method = c("fisher", "closed_form")))
  }else{
  method <- match.arg(method)
  df <- as.data.frame(enrich_res)

  if (!all(c("GeneRatio", "BgRatio", "Count") %in% names(df))) {
    stop("Expected columns 'GeneRatio', 'BgRatio', and 'Count' not found.")
  }

  # Parse ratios safely
  parse_ratio_num <- function(x, idx) {
    # idx=1 numerator, idx=2 denominator
    as.numeric(vapply(strsplit(x, "/", fixed = TRUE), function(z) z[[idx]], FUN.VALUE = character(1L)))
  }

  k <- as.numeric(df$Count)                                   # overlap in query ∩ term
  n <- parse_ratio_num(df$GeneRatio, 2L)                      # size of query
  M <- parse_ratio_num(df$BgRatio, 1L)                        # term size in background
  N <- parse_ratio_num(df$BgRatio, 2L)                        # background size

  # Sanity checks to catch inconsistencies early
  bad <- is.na(k) | is.na(n) | is.na(M) | is.na(N) | k > n | k > M | n > N | M > N
  if (any(bad)) {
    warning("Some rows have inconsistent counts; OR set to NA for those rows.")
  }

  # Compute ORs
  or <- rep(NA_real_, nrow(df))
  ci_lo <- rep(NA_real_, nrow(df))
  ci_hi <- rep(NA_real_, nrow(df))

  for (i in seq_len(nrow(df))) {
    if (bad[i]) next

    a <- k[i]
    b <- n[i] - k[i]
    c <- M[i] - k[i]
    d <- N[i] - M[i] - b

    if (any(c(a, b, c, d) < 0)) {
      next
    }

    if (method == "fisher") {
      # Why: fisher.test gives finite OR + exact CI, handles zeros better than closed-form
      mat <- matrix(c(a, b, c, d), nrow = 2,
                    dimnames = list(c("Query", "NotQuery"), c("InTerm", "NotInTerm")))
      ft <- suppressWarnings(fisher.test(mat, alternative = "greater"))
      or[i] <- unname(ft$estimate)
      ci_lo[i] <- ft$conf.int[1]
      ci_hi[i] <- ft$conf.int[2]
    } else {
      # Closed-form; may be Inf with zeros
      num <- a * d
      den <- b * c
      if (den == 0) {
        or[i] <- Inf
        ci_lo[i] <- NA_real_
        ci_hi[i] <- NA_real_
      } else {
        or[i] <- num / den
        ci_lo[i] <- NA_real_
        ci_hi[i] <- NA_real_
      }
    }
  }

  df$odds_ratio <- or
  df$or_ci_low  <- ci_lo
  df$or_ci_high <- ci_hi
  df
  }
}
# Run OR
file.name <- "04_popDTU-ORA/01_Processed-Data/06_celltype-fractions/02_popDTU-ORA-filtered-grouped-01_lineage.rds"
dtu.ora <- readRDS(file.name)
dtu.ora <-  add_enrichment_or(dtu.ora, method="fisher")
saveRDS(dtu.ora, file.name)
```

## 1.3 Group GO::Biological Process terms by similarity with revigo 

*We group similar GO terms with rrvgo (Sayols, 2023) to be able to fit a main figure*
```{r}
file.name <- "04_popDTU-ORA/01_Processed-Data/06_celltype-fractions/01_popDTU-ORA-clusterProfiler-GO-BP-01_lineage.rds"
dtu.ora <- readRDS( file.name)
# dtu.ora <- dtu.ora$lineage
# 1. Select 
dtu.ora <- get.result.ora(dtu.ora, fdr.thres = 0.05)
# 2. Group with revigo
rrvgo <- function ( df, db = "BP", p.value.thres = 0.05, sim.thres = 0.7 ){
  if(class(df) == "list"){ lapply(df, function(df) rrvgo(df, db, p.value.thres, sim.thres ))
  }else{
    print("Calculating the similarity matrix and reducing GO terms")
    if(!is.null(df)){
      if(nrow(df) > 1){
    # 0. Filter again by p.value
    df <- df[df$p.adjust < p.value.thres, ]
    # 1. Calculating the similarity matrix and reducing GO terms
    simMatrix <- calculateSimMatrix(df$ID, orgdb="org.Hs.eg.db", ont = db, method="Rel")
    # 2. Group terms based on similarity
    scores <- setNames(-log10(df$qvalue), df$ID) # in vignette they use q_value
    reducedTerms <- reduceSimMatrix(simMatrix, scores, threshold=sim.thres, orgdb="org.Hs.eg.db")
    # 3. Add to ORA df
    merge(df, reducedTerms, by.x = "ID", by.y = "go")
      }
    }
  }
}

rrvgo.l <- rrvgo( dtu.ora, db = "BP", p.value.thres = 0.05, sim.thres = 0.7)
names(rrvgo.l) <- names(dtu.ora)

#  3. Summarise df *Get N of GO terms per grouped term and relative fraction*
suppressPackageStartupMessages(require(dplyr))
count.rrvigo <- function(df) {
  if(class(df) == "list"){ lapply(df, count.rrvigo)
  }else{
    if(!is.null(df)){
    ## N of go_terms in each parent_term for each cell type and comparison
    n.go.terms.per.parent = df %>% group_by(parentTerm) %>% summarise(n_terms_parent = n_distinct(term))
    ## Total N of GO terms
    n.terms.total <- df %>% summarise(n_terms_total = n_distinct(term))
    ## Merge to calculate rel freq
    count.df = merge(n.go.terms.per.parent, n.terms.total)
    ## Relative N of terms
    count.df[["n_terms_parent_rel"]] <- count.df$n_terms_parent /  count.df$n_terms_total
    return(count.df)
    }
  }
}

count.rrvgo.l <- count.rrvigo(rrvgo.l)
# Merge
# By database
l <- lapply(count.rrvgo.l, function(ct) lapply(ct, function(c) lapply(c, function(set) plyr::ldply(set, .id = "GO_database")) ))
# By set
l <- lapply(l, function(ct) lapply(ct, function(c) plyr::ldply(c, .id = "set")) )
# By Condition
l <- lapply(l, function(ct) plyr::ldply(ct, .id = "Condition"))

# Save
file.name <- "04_popDTU-ORA/01_Processed-Data/06_celltype-fractions/03_popDTUDTU-ORA-revigo-grouped-terms-sim-0.7-01_lineage.rds"
saveRDS( l, file.name )
```
 
## 2. Supp Fig 13 

*Supplementary Fig. 13.* Pathway enrichment analysis of population-associated differential transcript usage (DTU) genes.Overrepresentation analysis (ORA) of population DTU genes identified across immune lineages under baseline, influenza A virus (IAV), and SARS-CoV-2 (COV) conditions. Gene sets include: (i) all DTU genes (DTU); (ii) DTU genes overlapping with differentially expressed genes (DTU & DGE); and (iii) DTU & DGE genes further stratified as upregulated in Africans (DTU & DGE AFR) or upregulated in Europeans (DTU & DGE EUR). Gene Ontology (GO) biological process terms with FDR < 0.05 (Benjamini–Hochberg correction) are shown. GO terms were grouped using rrvgo4 under a similarity threshold of 0.7. Dot color intensity and size represent the fraction and absolute number of GO terms within each grouped category, respectively.


```{r}
file.name <- "04_popDTU-ORA/01_Processed-Data/06_celltype-fractions/03_popDTUDTU-ORA-revigo-grouped-terms-sim-0.7-01_lineage.rds" 
term.l <- readRDS(file.name)
df <- plyr::ldply(term.l, .id = "lineage")
set.rename.vec <-  c("DTU.all" = "DTU", "DTU.unique" = "DTU not DGE", "DTU.DGE" = "DTU-DGE", 
                     "DTU.DGE.Up" = "DTU-DGE EUR" , "DTU.DGE.Down" = "DTU-DGE AFR")

df$celltype <- factor(df$lineage, levels = lineage.df$lineage)
df$Condition <- factor(df$Condition, levels = names(condition.vec.short))
df$Condition <- plyr::revalue(df$Condition, condition.vec.short)

plt <- ggplot(df,  aes(x = set, y = parentTerm, color = lineage, fill = celltype)) + # y = reorder(term_name_short, -p_value))) +
    geom_point( shape = 21, aes(alpha = n_terms_parent_rel, size = n_terms_parent)) +
    facet_grid( lineage ~ Condition, space = "free", scales = "free",  drop = T, switch="y") +
  xlab("Gene set") +
    scale_size_continuous("N GO terms\nin grouped term", 
                          breaks = c(1, 5, 10, 15), labels = c("1", "<5", "<10", "<15"), range = c(1, 3)) +
    scale_alpha_continuous("% GO terms\nper grouped term ", range = c(0.2, 1)) +
    scale_fill_manual("Lineage", values = lineage.color.vec) +
  scale_color_manual("Lineage", values = lineage.color.vec) +
    scale_y_discrete(labels = function(x) stringr::str_trunc(x, 40), limits = rev) + 
    # scale_color_discrete(labels = set.rename.vec, drop=T) + # if you want all sets plotted, drop = F
  scale_x_discrete(labels = set.rename.vec, drop=T) +
  plot_theme +   
  theme(panel.grid = element_line(color = "grey", linewidth = 0.1), 
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), 
        axis.title.x = element_blank(), axis.title.y = element_blank()) + 
  guides(color = F, 
         size = guide_legend(override.aes = list(fill="black")),
         alpha=guide_legend( override.aes = list(label = "", size = 3, ncol=1, fill = "black")),
         fill = guide_legend(override.aes = list(size = 3, color = "white")))
plt

file.name <-"04_popDTU-ORA/02_Plots/06_celltype-fractions/04_03_popDTU-ORA-all-conditions-database-rrvgo-sim-0.7.pdf"
ggsave(file.name, plt, width = 8.5, height =7, units = "cm")
```

